# Network Data {.unnumbered}

```{r}
#| label: libraries
#| message: false
library(igraph)
```

## Basic Network Notation

Networks are commonly represented with an **adjacency matrix** or via an **edgelist**. If you are interested 
Below, we represent friendship relations between Bob, Ann, and Steve as a matrix and an edgelist.
```{r}
#| label: simple_struc

# adjacency matrix
A <- matrix(
    c(0, 1, 1, 1, 0, 1, 1, 1, 0),
    nrow = 3, ncol = 3, byrow = TRUE
)

rownames(A) <- colnames(A) <- c("Bob", "Ann", "Steve")
A
# edgelist
el <- matrix(c("Bob", "Ann", "Bob", "Steve", "Ann", "Steve"),
    nrow = 3, ncol = 2, byrow = TRUE
)
el
```

The adjacency matrix $A$ is symmetric, meaning that the relations are undirected, i.e. Bob is friends with
Ann and Ann is friends with Bob. In general, $A[i,j]=1$, if there is a relation between $i$ and $j$.
If $A[i,j]=1$ does not imply $A[j,i]=1$ then $A$ defines a directed network.

Once we have defined an edgelist or an adjacency matrix, we can turn them into `igraph` objects as follows.

```{r}
#| label: simple-graph
g1 <- graph_from_adjacency_matrix(A, mode = "undirected", diag = FALSE)

g2 <- graph_from_edgelist(el, directed = FALSE)
# g1 and g2 are the same graph so only printing g1
g1
```

The printed summary shows some general descriptives of the graph.
The string "UN--" in the first line indicates that the network is *U*ndirected (*D* for directed graphs) and has a *N*ame attribute (we named the nodes Bob, Ann, and Steve). The third and forth character are *W*, if there is a edge weight attribute, and *B* if the network is bipartite (there exists a node attribute "type"). The following number indicate the number of nodes and edges.
The second line lists all graph, node and edge variables. Here, we only have a node attribute "name". 

The conversion from edgelist/adjacency matrix into an igraph object is quite straightforward. The only difficulty is setting the parameters correctly (Is the network directed or not?), especially for edgelists where it may not immediately be obvious if the network is directed or not.
