# Introduction
Most network analytic tasks are fairly straightforward to do in R. But when it comes to visualizing
networks, R may lack behind some standalone software tools. Not because it is not possible to produce nice
figures, but rather because it requires some time to obtain pleasing results.
Just take a look at the default output when plotting a network with the `plot()` function.
```{r}
#| label: silent_igraph
#| include: false

library(igraph)
library(patchwork)
```

```{r}
#| label: plot_ugly
library(networkdata)
data("got")

gotS1 <- got[[1]]
plot(gotS1)
```

It is definitely possible to produce nice figures with the igraph package (Check out [this](https://kateto.net/networks-r-igraph) wonderful tutorial), yet it may take some time
to familiarize yourself with the syntax. Additionally, most of the layout algorithms of `igraph` 
are non-deterministic. This means that running the same plot call twice may produce different results.

In this part, you will learn the basics of `ggraph`, the "`ggplot2` of networks", together with
the `graphlayouts` package, which introduces additional useful layout algorithms to R. 
Arguably, using `ggraph` is not really easier than `igraph`. But once the underlying 
principle of the *grammar of graphics* is understood, you'll see that it is actually quite intuitive to work with. 

## Required libraries

To run all the code in this tutorial, you need to install and load several packages.
```{r}
#| label: install_libraries
#| eval: false
install.packages(c("igraph", "graphlayouts", "ggraph", "ggforce"))
devtools::install_github("schochastics/networkdata")
```

Make sure you have at least the version given below. Some of the examples may not be backward compatible.

```{r}
#| label: pkg_version_check
packageVersion("igraph")
packageVersion("graphlayouts")
packageVersion("ggraph")
packageVersion("networkdata")
packageVersion("ggforce")
```

`igraph` is mostly used for its data structures and `graphlayouts` and `ggraph` for visualizations.
The `networkdata` package contains a huge amount of example network data that always comes in handy for learning new visualization techniques.

```{r}
#| label: libraries
library(igraph)
library(ggraph)
library(graphlayouts)
library(ggforce)
```
