# Basics of tidygraph

This chapter walks through the foundation of tidygraph before any network descriptors come into play: how a graph is stored as a pair of tables in a `tbl_graph`, how `activate()` decides which of the two tables a verb sees, how the standard `dplyr` verbs and joins carry over, and what the tidygraph-specific `morph()` family adds on top. The Florentine marriage network from the Descriptive part is reused throughout so that each idiom can be compared against the `igraph` equivalent you have already seen.

## Packages Needed for this Chapter

```{r}
#| label: libraries
#| message: false
library(tidygraph)
library(networkdata)
library(dplyr)
```

```{r}
#| label: libraries-silent
#| include: false
library(ggraph)
library(graphlayouts)
```

## Graph Structures

We will use the Florentine Families marriage network as a running example. The dataset ships as an `igraph` object and converts to a `tbl_graph` with `as_tbl_graph()`.

```{r}
#| label: load-flo
data("flo_marriage")
flo_tidy <- as_tbl_graph(flo_marriage)
flo_tidy
```

Printing a `tbl_graph` surfaces the two tables it wraps: a node data frame and an edge data frame. The header tells you which of the two is currently active.

```{r}
#| label: class-flo
class(flo_tidy)
```

The class `tbl_graph` inherits from `igraph`, so anything written against `igraph` accepts a `tbl_graph` unchanged. No conversion is needed to go back to an `igraph` workflow.

If you need to build a network from scratch, `tbl_graph()` takes the two data frames directly and is essentially the tidy counterpart of `graph_from_data_frame()`. For the common random graph generators, the `create_*()` family produces deterministic graphs (lattices, stars, rings) and `play_*()` produces stochastic ones (Erdős–Rényi, preferential attachment, island models).

```{r}
#| label: example-random-graphs
create_ring(5)
play_gnp(10, 0.3)
play_barabasi_albert(10,2)
```

## Standard Verbs

The tidy idiom revolves around a small set of verbs like `mutate()`, `select()`, `filter()`, `summarise()` and joins that transform a data frame. The awkwardness for networks is that a `tbl_graph` wraps two tables, so a verb has to know which one to operate on. `tidygraph` solves this with an explicit pointer: whichever table is **active** is the one the verb sees. The pointer is set with `activate()`. The default is `"nodes"`, which is why the example above printed the node table on top.

```{r}
#| label: activate-edges
flo_tidy |> activate("edges")
```

Switching the active table lets you attach new edge attributes the same way you would add a new column to any tibble. A useful trick when you need information from the other table is `.N()` and `.E()`: `.N()` returns the node table while edges are active, and `.E()` returns the edge table while nodes are active. The example below uses `.N()` to tag each edge according to whether one of its endpoints is the Medici family.

```{r}
#| label: tag-medici-edges
flo_medi <- flo_tidy |>
    activate("edges") |>
    mutate(to_medici = (.N()$name[from] == "Medici" | .N()$name[to] == "Medici"))
```

The new attribute is immediately available for visualisation, as shown in @fig-connect-medici.

```{r}
#| label: fig-connect-medici
#| fig-cap: "Florentine marriage network with edges coloured by whether they touch the Medici family."
ggraph(flo_medi, "stress") +
    geom_edge_link0(aes(edge_color = to_medici)) +
    geom_node_point(shape = 21, size = 10, fill = "grey66") +
    geom_node_text(aes(label = name)) +
    theme_graph()
```

`filter()` behaves the way the table-based intuition suggests, with one asymmetry worth remembering. Filtering on nodes also drops every edge that was incident to a removed node; filtering on edges leaves the node set intact. In @fig-filter-medici, only the edges touching the Medici survive, while the now-isolated families remain in the plot.

```{r}
#| label: fig-filter-medici
#| fig-cap: "Result of filtering the edge table to ties incident to the Medici. Nodes untouched by any surviving edge remain in the graph as isolates."
flo_medi |>
    activate("edges") |>
    filter(to_medici) |>
    ggraph("stress", bbox = 10) +
    geom_edge_link0(edge_color = "black") +
    geom_node_point(shape = 21, size = 10, fill = "grey66") +
    geom_node_text(aes(label = name)) +
    theme_graph()
```

## Joins

Networks are sometimes more naturally described as several smaller networks on overlapping actors. The classic Florentine study, for instance, actually records two relations among the same families: marriage ties and business ties. tidygraph offers a handful of verbs for combining such graphs and for pulling in external attribute tables.

`graph_join()` merges two `tbl_graph` objects on a shared node key (by default the `name` column). Nodes that appear in both graphs are identified; edges from both graphs are kept. The `networkdata` package ships `flo_business`, the second Padgett relation over the same 16 families, which we load now and turn into a `tbl_graph`.

```{r}
#| label: load-flo-business
data("flo_business")
flo_business_tidy <- as_tbl_graph(flo_business)
```

Before merging, we tag each graph's edges with the relation they come from, so the combined plot in @fig-graph-join can colour edges by their source network.

```{r}
#| label: fig-graph-join
#| fig-cap: "Marriage and business ties among the Florentine families merged with `graph_join()`. Edges are coloured by the relation they originated from."
marriage_rel <- flo_tidy |>
    activate("edges") |>
    mutate(relation = "marriage")

business_rel <- flo_business_tidy |>
    activate("edges") |>
    mutate(relation = "business")

marriage_rel |>
    graph_join(business_rel, by = "name") |>
    ggraph("stress") +
    geom_edge_parallel0(aes(edge_color = relation)) +
    geom_node_point(shape = 21, size = 10, fill = "grey66") +
    geom_node_text(aes(label = name)) +
    theme_graph()
```

When the goal is instead to keep two networks side by side, for example to lay them out together without merging their node sets, `bind_graphs()` returns the disjoint union. Stress layout then places each component independently, as @fig-bind-graphs shows.

```{r}
#| label: fig-bind-graphs
#| fig-cap: "Disjoint union of the Florentine marriage and business networks produced by `bind_graphs()`. Each family appears twice and the two components are laid out independently."
bind_graphs(
    flo_tidy |> activate("nodes") |> mutate(source = "marriage"),
    flo_business_tidy |> activate("nodes") |> mutate(source = "business")
) |>
    ggraph("kk") +
    geom_edge_link0() +
    geom_node_point(aes(fill = source), shape = 21, size = 8) +
    geom_node_text(aes(label = name), size = 3) +
    theme_graph()

```

Two more verbs, `bind_nodes()` and `bind_edges()`, append rows to the active table and are the tidygraph analogues of `dplyr::bind_rows()`. They are useful when new nodes or edges arrive as a plain data frame.

Finally, because each table inside a `tbl_graph` is a tibble, any standard `dplyr` join works once that table is activated. This is the path of least resistance for attaching an external attribute table keyed on the node name:

```{r}
#| label: dplyr-join
family_estate <- data.frame(
    name   = c("Medici", "Strozzi", "Peruzzi", "Guadagni"),
    estate = c(15, 14, 2, 2)
)

flo_tidy |>
    activate("nodes") |>
    left_join(family_estate, by = "name")
```

Nodes without a match simply get `NA` for the new column, exactly as with an ordinary tibble join.

## Special Graph Verbs

Beyond `activate()`, tidygraph introduces a small set of verbs that have no direct analogue in `dplyr` because they only make sense on graphs. The most important is `morph()`. A morph temporarily reshapes a graph into a different representation, for example the connected components, the shortest path between two nodes, the line graph, or the minimum spanning tree, without committing to that change. Verbs chained after `morph()` run on the morphed representation, and `unmorph()` returns to the original graph, propagating any attributes that were added inside the morph back onto the original nodes or edges.

As a first example, we attach the size of each connected component to every node. Inside the `to_components` morph the graph is effectively split into one sub-graph per component, so `graph_order()` evaluates separately for each.

```{r}
#| label: morph-components
flo_tidy |>
    activate("nodes") |>
    morph(to_components) |>
    mutate(component_size = graph_order()) |>
    unmorph()
```

The Pucci family, the lone isolate, now carries `component_size = 1`; every other node carries the size of the giant component.

If you do want to commit to the morphed graph rather than return to the original, use `convert()` instead of the `morph()` / `unmorph()` pair. The next example extracts the shortest path between the Medici and the Pazzi and keeps only that path as the new graph.

```{r}
#| label: convert-shortest-path
medici_id <- which(igraph::V(flo_tidy)$name == "Medici")
pazzi_id  <- which(igraph::V(flo_tidy)$name == "Pazzi")

flo_tidy |>
    convert(to_shortest_path, medici_id, pazzi_id)
```

A related verb, `crystallise()`, is worth knowing for morphs that naturally return multiple graphs, for instance `to_components` again, or `to_local_neighborhood` called on several nodes at once. It materialises the morph as a tibble with one row per sub-graph, which is convenient when you want to iterate.

The catalogue of morphers is large: `to_undirected`, `to_directed`, `to_simple`, `to_subgraph`, `to_contracted`, `to_complement`, `to_minimum_spanning_tree`, `to_dominator_tree`, `to_linegraph`, `to_subcomponent`, and more. Rather than enumerate them all, the recommendation is to browse `?morphers` once so you know what is available and reach for it when the shape of your problem calls for a transformed graph.
