# Descriptive Network Analysis

This chapter revisits the descriptors introduced in the Descriptive part — centrality, community structure, and the assortment of node- and edge-level measures — through tidygraph. The point is not to teach the descriptors again but to show how the same computations translate into pipeable `mutate()` calls on an activated node or edge table. Cross-references to the original treatment in [Centrality](../descriptive/centrality-basic.qmd) and [Cohesive Subgroups](../descriptive/clustering.qmd) are given where they help.

## Packages Needed for this Chapter

```{r}
#| label: libraries
#| message: false
library(tidygraph)
library(networkdata)
```

```{r}
#| label: libraries-silent
#| include: false
library(ggraph)
library(graphlayouts)
```

```{r}
#| label: load-flo
data("flo_marriage")
flo_tidy <- as_tbl_graph(flo_marriage)
```

## Centrality

Centrality indices live in the `centrality_*()` function family. The family covers everything implemented in `igraph` — degree, closeness, betweenness, eigenvector, PageRank, Katz, hub and authority scores — together with the richer catalogue exposed by the `netrankr` package. Every function in the family shares the same calling convention: no graph argument, no vertex argument, just the name. Each one reads the currently active graph implicitly and returns a vector the length of the active table, which is what makes them slot straight into `mutate()` alongside any other derived column. @fig-centrality-example combines degree and betweenness in a single pipeline and maps them to node size and fill.

```{r}
#| label: fig-centrality-example
#| fig-cap: "Florentine marriage network with node size scaled to degree centrality and fill scaled to betweenness centrality."
flo_tidy |>
    activate("nodes") |>
    mutate(
        degree      = centrality_degree(),
        betweenness = centrality_betweenness()
    ) |>
    ggraph("stress", bbox = 10) +
    geom_edge_link0(edge_color = "black") +
    geom_node_point(shape = 21, aes(size = degree, fill = betweenness)) +
    geom_node_text(aes(label = name)) +
    scale_fill_gradient(low = "#104E8B", high = "#CD2626") +
    scale_size(range = c(4, 10)) +
    theme_graph()
```

A practical tip: because every function starts with the same prefix, typing `centrality_` in most IDEs and pressing Tab is the fastest way to browse what is available without consulting the help pages first.

## Clustering

Community detection algorithms are exposed through the parallel `group_*()` family: `group_louvain`, `group_infomap`, `group_edge_betweenness`, `group_walktrap`, `group_leading_eigen`, `group_fast_greedy`, `group_label_prop`, `group_spinglass`, `group_components`, `group_biconnected_component`, and a few more. Each returns an integer membership vector the length of the node table, which means the idiomatic usage is `mutate(community = as.factor(group_*()))` — coerce to factor so the result becomes a natural input to a `fill` or `colour` aesthetic.

We illustrate the pattern on a random graph with planted community structure, generated with `play_islands()` (the tidygraph equivalent of `igraph::sample_islands()`). The Louvain partition recovered on that graph is shown in @fig-clustering-louvain.

```{r}
#| label: fig-clustering-louvain
#| fig-cap: "Louvain community detection applied to a graph with planted community structure generated by `play_islands()`. Node fill encodes the recovered community membership."
play_islands(4, 12, 0.8, 4) |>
    mutate(community = as.factor(group_louvain())) |>
    ggraph(layout = "stress") +
    geom_edge_link0() +
    geom_node_point(aes(fill = community), shape = 21, size = 6) +
    theme_graph()
```

A frequent follow-up is to colour edges by community — one colour when an edge stays inside a cluster, another when it bridges two clusters. The way to do this is to propagate the node-level community attribute into the edge table using `.N()[from]` and `.N()[to]`, which look up the community of each edge's endpoints. If they match, the edge is intra-community; otherwise, we mark it as a bridge. @fig-clustering-bridges shows the resulting picture.

```{r}
#| label: fig-clustering-bridges
#| fig-cap: "Same graph as in @fig-clustering-louvain, with edges additionally coloured to distinguish intra-community ties from bridges between communities."
play_islands(4, 12, 0.8, 4) |>
    mutate(community = as.factor(group_louvain())) |>
    activate("edges") |>
    mutate(community = as.factor(ifelse(
        .N()$community[from] == .N()$community[to],
        .N()$community[from],
        5
    ))) |>
    ggraph(layout = "stress") +
    geom_edge_link0(aes(edge_colour = community), show.legend = FALSE) +
    geom_node_point(aes(fill = community), shape = 21, size = 6) +
    scale_fill_brewer(palette = "Set3") +
    scale_edge_color_brewer(palette = "Set3") +
    theme_graph(background = "grey88")
```

## Other node or edge level functions

Beyond centrality and grouping, tidygraph harmonises a long list of node- and edge-level measures behind two mirror-image function families: `node_*()` returns a vector of length `vcount()`, and `edge_*()` returns a vector of length `ecount()`. Which family you reach for is determined by which table you have activated. The families cover distance (`node_distance_to`, `node_distance_from`, `node_eccentricity`), local structure (`node_triangles`, `node_coreness`, `node_local_transitivity`), and tests on edges (`edge_is_loop`, `edge_is_multiple`, `edge_is_mutual`, `edge_is_incident`), among many others. As with the centrality and group families, typing `node_` or `edge_` at the prompt and pressing Tab gives the full list.

The example below combines one of each: on nodes, the geodesic distance to the Medici family; on edges, a flag indicating whether the edge is incident to the Medici. Activating each table in turn keeps both computations in a single pipeline, and the result is shown in @fig-distance-medici.

```{r}
#| label: fig-distance-medici
#| fig-cap: "Florentine marriage network with node fill encoding the geodesic distance to the Medici family and edge colour flagging ties incident to the Medici."
flo_tidy |>
    activate("nodes") |>
    mutate(dist2Medici = node_distance_to(nodes = 9)) |>
    activate("edges") |>
    mutate(edge2Medici = edge_is_incident(9)) |>
    ggraph("stress") +
    geom_edge_link0(aes(edge_color = edge2Medici)) +
    geom_node_point(aes(fill = dist2Medici), size = 9, shape = 21) +
    theme_graph()
```

The node id `9` refers to the Medici, which happens to be the ninth row of the node table. Passing the index directly is the quickest way when you already know the position; for a name-based lookup you can equally write `nodes = which(.N()$name == "Medici")`.
