# Cohesive Subgroups {#sec-cohesive-subgroups}

In this chapter, we explore cohesive subgroups within networks, focusing on cliques, community detection, blockmodeling, and core-periphery structures. Each concept offers a different way to examine how nodes cluster based on connectivity, from tightly-knit cliques, where every member is connected, to broader communities grouped by denser internal links. Blockmodeling abstracts these patterns into roles and relationships, while core-periphery structures reveal hierarchical organization with central and peripheral actors.


## Packages Needed for this Chapter

```{r}
#| label: libraries
#| message: false
library(igraph)
library(networkdata)
```

```{r}
#| label: libraries_silent
#| include: false
library(ggraph)
library(patchwork)
```

## Cliques 

A *clique* in a network is a set of nodes that form a complete subnetwork within a network (called a complete *subgraph*). A *maximal clique* is a clique that cannot be extended to a bigger clique by adding more nodes to it.

The `networkdata` package includes a small example graph called `clique_graph` which contains several maximal cliques of different sizes. We will use this graph to illustrate how to find cliques in a network.
```{r}
#| label: load-clique-graph
data("clique_graph")
```

```{r}
#| label: plot-clique-graph-blank
#| echo: false
#| fig-cap: "The example network used to illustrate cliques and k-core decomposition."
ggraph(clique_graph, "stress") +
  geom_edge_link0(edge_color = "grey66") +
  geom_node_point(shape = 21, size = 8, fill = "grey25") +
  theme_void()
```

All maximal cliques can be calculated with `max_cliques()`. This is only feasible for fairly small networks given the complexity of the task. The `min` parameter can be used to set a minimum size of discovered cliques. Below, we are only interested to find maximal cliques of size 3 or more.

```{r}
#| label: calc_max_cliques
# only return cliques with three or more nodes
cl <- max_cliques(clique_graph, min = 3)
cl
```

The figure shows the network and the uncovered maximal cliques.

```{r}
#| label: plot-clique-graph
#| echo: false
#| fig-cap: "Maximal cliques of size 3 or more highlighted in the example network."

xy <- graphlayouts::layout_with_stress(clique_graph)

cl_df <- as.data.frame(do.call(
  "rbind",
  lapply(seq_along(cl), function(x) {
    cbind(xy[cl[[x]], ], x)
  })
))

ggraph(clique_graph, "stress") +
  geom_edge_link0(edge_color = "grey66") +
  geom_node_point(shape = 21, size = 8, fill = "grey25") +
  ggforce::geom_mark_hull(
    data = cl_df,
    aes(V1, V2, fill = as.factor(x), group = x),
    show.legend = FALSE
  ) +
  scale_fill_manual(
    values = c(
      "#E69F00",
      "#000000",
      "#56B4E9",
      "#009E73",
      "#F0E442",
      "#0072B2",
      "#D55E00",
      "#CC79A7",
      "#666666"
    )
  ) +
  theme_void()

```

Related to cliques is the **k-core decomposition** of a network. A k-core is a subgraph in which every node has at least k neighbors within the subgraph. A k-core is thus a relaxed version of a clique.  
The function `coreness()` can be used to calculate the k-core membership for each node.
```{r}
#| label: kcore
kcore <- coreness(clique_graph)
kcore
```


```{r}
#| label: kcore_plot
#| echo: false
#| fig-cap: "K-core decomposition of the example network. Nested hulls show the 2-core, 3-core, and 4-core."

cl_df <- as.data.frame(do.call(
  "rbind",
  lapply(sort(unique(kcore))[c(2, 3, 4)], function(x) {
    cbind(xy[kcore >= x, ], x)
  })
))

ggraph(clique_graph, "stress") +
  geom_edge_link0(edge_color = "grey66") +
  geom_node_point(shape = 21, size = 8, fill = "grey25") +
  ggforce::geom_mark_hull(
    data = cl_df,
    aes(V1, V2, fill = as.factor(x), group = x),
    show.legend = FALSE
  ) +
  scale_fill_manual(values = c("red", "blue", "green")) +
  theme_void()

```

Cliques are the prototypical and most strict definition of a cohesive subgroup in a network. In empirical networks, however, we rarely encounter situations where we can partition the whole network into a set of cliques. Real networks are messy: ties are missing, groups overlap, and boundaries are fuzzy. We therefore need methods that find approximate groupings based on the overall pattern of connections rather than requiring every possible tie to be present.

## Community Detection

Unlike cliques, which have a precise mathematical definition, the notion of a "community" or "cluster" in a network is inherently fuzzy. The general intuition, however, is straightforward: a community is a group of nodes that are more densely connected to each other than to the rest of the network. But there is no universally accepted formal definition of what this actually means. How dense is dense enough? How sparse must the connections to outsiders be? Different algorithms operationalise this intuition in different ways, which is why they can produce different results on the same network. This is not a flaw but a reflection of the fact that "community" is a flexible concept whose meaning depends on the context of the analysis.

The example below shows a network with a visible and intuitive cluster structure.

```{r}
#| label: clustered-graph
#| echo: false
#| fig-cap: "A network with a clearly visible cluster structure."

n1 <- 5
n2 <- 20
set.seed(1234)
g <- sample_islands(n1, n2, 0.9, 5)
g <- simplify(g)
V(g)$grp <- rep(LETTERS[1:n1], each = n2)
ggraph(g, "stress") +
  geom_edge_link0(edge_linewidth = 0.2, edge_color = "grey66") +
  geom_node_point(shape = 21, size = 5, fill = "grey66") +
  theme_void()
```

In contrast, the network below does not really seem to have any well defined cluster structure on first sight.

```{r}
#| label: random-graph
#| echo: false
#| fig-cap: "A network without obvious cluster structure."
n1 <- 5
n2 <- 20

set.seed(1234)
g <- sample_islands(n1, n2, 0.25, 15)
g <- simplify(g)
V(g)$grp <- rep(LETTERS[1:n1], each = n2)
ggraph(g, "stress") +
  geom_edge_link0(edge_linewidth = 0.2, edge_color = "grey66") +
  geom_node_point(shape = 21, size = 5, fill = "grey66", show.legend = FALSE) +
  theme_void()

```

The `igraph` package includes a number of different clustering algorithms as shown below.

```{r}
#| label: cluster_alg
#| echo: FALSE
algs <- as.character(lsf.str("package:igraph"))
algs[stringr::str_detect(algs, "cluster_")]
```

Most of these algorithms are based on "modularity maximization". The idea behind modularity is to compare the observed number of edges within groups to what we would expect if edges were placed at random while preserving each node's degree. Formally, modularity is defined as

$$
Q = \frac{1}{2m} \sum_{ij} \left[ A_{ij} - \frac{k_i k_j}{2m} \right] \delta(c_i, c_j)
$$

where $m$ is the total number of edges, $A_{ij}$ is the adjacency matrix entry (1 if nodes $i$ and $j$ are connected, 0 otherwise), $k_i$ is the degree of node $i$, and $\delta(c_i, c_j)$ equals 1 if nodes $i$ and $j$ are assigned to the same community and 0 otherwise. The term $\frac{k_i k_j}{2m}$ is the expected number of edges between $i$ and $j$ under a null model that preserves degree. Modularity values typically range from around $-0.5$ to $1$, where values close to zero indicate no more within-group edges than expected by chance, and higher values indicate stronger community structure.

The workflow of a cluster analysis is always the same, independent from the chosen method. We illustrate the workflow using the infamous karate club network. This network was collected by Wayne Zachary in the 1970s and it captures friendships among 34 members of a university karate club. During the study, a conflict between the club's instructor and the administrator led to the club splitting into two factions, making it a classic benchmark for community detection.
```{r}
#| label: karate
data("karate")
```

```{r}
#| label: karate-plot
#| echo: false
#| fig-cap: "Zachary's karate club network."
ggraph(karate, "stress") +
  geom_edge_link0(edge_color = "grey66") +
  geom_node_point(shape = 21, size = 5, fill = "grey66") +
  theme_void() +
  coord_equal()
```

First, we pick an algorithm and run it on the network. Here, we use the Louvain method for modularity maximization via `cluster_louvain()`. The result is a community object from which we can extract a membership vector (which community each node belongs to) and a list of communities (which nodes belong to each community).

```{r}
#| label: cluster_ex
## compute clustering
clu <- cluster_louvain(karate)

## cluster membership vector
mem <- membership(clu)
mem

## clusters as list
com <- communities(clu)
com
```

Finding the partition that maximises modularity is a computationally hard problem (NP-hard, to be precise). No algorithm can guarantee finding the global optimum in reasonable time, so all practical methods rely on heuristics. Different heuristics explore the space of possible partitions in different ways, which means they can arrive at different solutions on the same network. To see how much results vary, we run several algorithms and compare their modularity scores.

```{r}
#| label: karate-cluster
imc <- cluster_infomap(karate)
lec <- cluster_leading_eigen(karate)
loc <- cluster_louvain(karate)
sgc <- cluster_spinglass(karate)
wtc <- cluster_walktrap(karate)
scores <- c(
  infomap = modularity(karate, membership(imc)),
  eigen = modularity(karate, membership(lec)),
  louvain = modularity(karate, membership(loc)),
  spinglass = modularity(karate, membership(sgc)),
  walk = modularity(karate, membership(wtc))
)
scores
```

For the karate network, `cluster_spinglass()` produces the highest modularity score.
In practice, the choice of algorithm depends on more than just modularity. `cluster_louvain()` scales well to large networks, while `cluster_spinglass()` tends to produce high-quality results but is only feasible for smaller networks.

The corresponding clustering is shown below.

```{r}
#| label: karate-plot-clu
#| echo: false
#| fig-cap: "Karate club network coloured by community membership (spinglass algorithm)."
V(karate)$clu <- membership(sgc)
ggraph(karate, "stress") +
  geom_edge_link0(edge_color = "grey66") +
  geom_node_point(
    shape = 21,
    size = 5,
    aes(fill = as.factor(clu)),
    show.legend = FALSE
  ) +
  theme_void() +
  coord_equal()
```

Modularity maximization is still widely considered as the state-of-the-art clustering method
for networks. There are, however, some technical shortcomings that one should be aware of.
One of those is the so called "resolution limit". When modularity is being maximized, it can happen
that smaller clusters are merged together to form bigger clusters. The prime example is the graph that
consists of cliques connected in a ring.

```{r}
#| label: prepare-Kn-graph
#| echo: false
n1 <- 5
n2 <- 50
A <- matrix(1, n1, n1)
lst <- vector("list", n2)
lst <- lapply(lst, function(x) A)
AA <- Matrix::bdiag(lst)
for (i in 1:(n2 - 1)) {
  AA[i * n1, i * n1 + 1] <- AA[i * n1 + 1, i * n1] <- 1
}
AA[1, n1 * n2] <- AA[n1 * n2, 1] <- 1
K50 <- graph_from_adjacency_matrix(AA, "undirected", diag = FALSE)
```

The figure below shows such a graph, consisting of 50 cliques of size 5. 

```{r}
#| label: plot-K50-blank
#| echo: false
#| fig-cap: "A ring of 50 cliques of size 5, used to illustrate the resolution limit."
#| width: 8
#| height: 8
ggraph(K50, "stress") +
  geom_edge_link0(edge_linewidth = 0.6, edge_color = "grey66") +
  geom_node_point(shape = 21, fill = "grey66", size = 2, show.legend = FALSE) +
  theme_void() +
  coord_fixed()
```

Intuitively, any clustering method should return a cluster for each clique.

```{r}
#| label: clu-louvain-K50
clu_louvain <- cluster_louvain(K50)
table(membership(clu_louvain))

```

A clustering algorithm that fixes this issue is the Leiden algorithm. It is available in `igraph` via `cluster_leiden()` and can be used as a drop-in replacement for `cluster_louvain()`. When used with the CPM (Constant Potts Model) objective function, it avoids the resolution limit entirely.

```{r}
#| label: clu-leiden-K50
clu_leiden <- cluster_leiden(
  K50,
  objective_function = "CPM",
  resolution_parameter = 0.5
)
table(membership(clu_leiden))

```

The figure below shows the clusters computed with the Louvain method in grey and the Leiden method in red.

```{r}
#| label: plot-K50-clu
#| echo: false
#| fig-cap: "Louvain clusters (grey) vs. Leiden clusters (red) on the ring-of-cliques network."
#| width: 8
#| height: 8
V(K50)$louvain <- membership(clu_louvain)
V(K50)$leiden <- membership(clu_leiden)
ggraph(K50, "stress") +
  geom_edge_link0(edge_linewidth = 0.6, edge_color = "grey66") +
  geom_node_point(
    shape = 21,
    fill = "grey66",
    size = 1.5,
    show.legend = FALSE
  ) +
  ggforce::geom_mark_ellipse(
    aes(x, y, group = louvain),
    expand = unit(4.5, "pt"),
    fill = "black",
    alpha = 0.25
  ) +
  ggforce::geom_mark_circle(
    aes(x, y, group = leiden),
    expand = unit(4.5, "pt"),
    col = "firebrick3",
    size = 1
  ) +
  theme_void() +
  coord_fixed()

```

If you are interested in the technical details of the Leiden method, check out the [original paper](https://www.nature.com/articles/s41598-019-41695-z). 

The `netUtils` package includes the function `sample_lfr()` which implements the well-known [Lancichinetti–Fortunato–Radicchi benchmark algorithm](https://en.wikipedia.org/wiki/Lancichinetti%E2%80%93Fortunato%E2%80%93Radicchi_benchmark) to generate artificial networks with a priori known communities and they can be used to compare different community detection methods.

### Choosing the number of communities

Some algorithms, like Louvain and Leiden, determine the number of communities automatically as part of the optimisation. Others, such as `cluster_spinglass()`, require the number to be specified upfront. When it is not known in advance, a common strategy is to run the algorithm for a range of values of $k$ and compare the resulting modularity scores. The value of $k$ that produces the highest modularity is then selected. However, modularity tends to increase with $k$ up to a point and then plateau or decrease, so the "best" $k$ is typically where the curve levels off rather than its absolute maximum.

```{r}
#| label: choose-k
scores_k <- sapply(2:8, function(k) {
  cl <- cluster_spinglass(karate, spins = k)
  modularity(karate, membership(cl))
})
names(scores_k) <- 2:8
round(scores_k, 3)
```

In practice, the number of communities should also be guided by substantive knowledge about the network. A purely data-driven choice of $k$ can lead to splits that are statistically optimal but socially meaningless.

### Overlapping communities

All methods discussed so far produce a hard partition: each node is assigned to exactly one community. In many social networks, however, people belong to multiple groups simultaneously. A person may be part of their work team, a sports club, and a circle of school friends, and these memberships can overlap. Hard partitions force a choice that may not reflect the actual social structure.

One way to get a sense of potential overlap is to examine which nodes participate in multiple maximal cliques. The `clique_vertex_mat()` function from `netUtils` returns a matrix indicating which nodes belong to which cliques.

```{r}
#| label: clique-vertex-mat
cvm <- netUtils::clique_vertex_mat(karate)
## only consider cliques of size 3 or more
cvm <- cvm[rowSums(cvm) >= 3, ]
## nodes that appear in more than one clique
overlap <- colSums(cvm) > 1
V(karate)$name[overlap]
```

Nodes that appear in multiple cliques are natural candidates for overlapping community membership. Dedicated methods for overlapping community detection exist (such as clique percolation), but well-maintained R implementations are scarce. In practice, ego networks (see @sec-ego) offer another perspective on overlapping membership, since each individual's personal network naturally captures the multiple social circles they participate in.

## Blockmodeling

Community detection focuses on finding groups of nodes that are densely connected internally. Blockmodeling takes a different perspective: it groups nodes based on *patterns* of connections, not just density. Two nodes belong to the same block if they relate to other blocks in similar ways, regardless of whether they are connected to each other. This makes blockmodeling particularly useful for role analysis in sociology and organisational studies.

There are several packages that implement different kinds of blockmodels. The most basic deterministic approaches are implemented in the package `blockmodeling`.

```{r}
#| label: lib-blockmodel
#| message: false
library(blockmodeling)
```

We illustrate blockmodeling on a random network with 3 dense blocks of size 20.

```{r}
#| label: random-graph2
#| echo: false
#| fig-cap: "A random network with three dense blocks of 20 nodes each."
n1 <- 3
n2 <- 20

set.seed(1234)
g <- sample_islands(n1, n2, 0.75, 5)
g <- simplify(g)
V(g)$grp <- rep(LETTERS[1:n1], each = n2)
ggraph(g, "stress") +
  geom_edge_link0(edge_linewidth = 0.2, edge_color = "grey66") +
  geom_node_point(shape = 21, size = 5, fill = "grey66", show.legend = FALSE) +
  theme_void()

```

Compared to community detection, blockmodeling requires us to specify more parameters. We need to define the number of blocks, the type of approach (binary or valued), and what block types we expect to find.

```{r}
#| label: calc-rnd-block
A <- as_adj(g)
## define expected block structure: complete on diagonal, null off-diagonal
blk <- matrix(
  c(
    "com", "nul", "nul",
    "nul", "com", "nul",
    "nul", "nul", "com"
  ),
  nrow = 3
)
blk
res <- optRandomParC(
  M = A,
  k = 3,               ## number of blocks
  approaches = "bin",   ## binary blockmodel
  blocks = blk,         ## expected block structure
  rep = 5,              ## random starting partitions
  mingr = 20,           ## min block size
  maxgr = 20            ## max block size
)
```

The result can be accessed with `clu()`.
```{r}
clu(res)
```

Note that blockmodeling is computationally expensive and best suited for small networks. The `blocks` parameter controls what types of blocks are allowed: "com" (complete) means all ties present within, "nul" (null) means no ties. Consult the package documentation for more block types. For signed networks, a dedicated blockmodeling approach is available in the `signnet` package (see @sec-signed-networks).

## Core-Periphery

Community detection aims to find clusters or groups of nodes within the network that are more densely connected to each other than to nodes outside the group. The core-periphery structure, in contrast, posits that a network is organized into a densely connected core and a sparsely connected periphery. The core consists of central nodes that are highly connected to each other and also to peripheral nodes. Peripheral nodes, on the other hand, have fewer connections and are mostly connected to nodes in the core rather than to each other. 

The `netUtils` package includes a function `core_periphery()` which allows to fit a discrete core-periphery model to a network.

```{r}
#| label: load_netUtils
library(netUtils)
```

To illustrate the function, we construct a graph which has a perfect core-periphery structure, also known as a **split graph**.

```{r}
#| label: split-graph
set.seed(1234)
sg <- split_graph(n = 50, p = 0.35, core = 0.3)
```

The created graph has 50 nodes and 30% of all nodes are in the core. The probability that a periphery node connects to a core node is 0.35.

```{r}
#| label: split-graph-img
#| echo: false
#| fig-cap: "A split graph with a perfect core-periphery structure. Left: network layout with core and periphery nodes. Right: adjacency matrix showing the block pattern."
V(sg)$type <- c("periphery", "core")[V(sg)$core + 1]
p1 <- ggraph(sg, "stress") +
  geom_edge_link0(edge_color = "grey66", edge_linewidth = 0.1) +
  geom_node_point(shape = 21, size = 5, aes(fill = type)) +
  theme_graph() +
  theme(legend.position = "bottom")

E(sg)$col <- apply(
  matrix(V(sg)$core[as_edgelist(sg, names = FALSE)], ncol = 2),
  1,
  all
)
E(sg)$type <- c("core - periphery", "core - core")[E(sg)$col + 1]
p2 <- ggraph(sg, "matrix") +
  geom_edge_point(mirror = TRUE, aes(edge_color = type)) +
  coord_fixed() +
  scale_y_reverse() +
  theme_graph() +
  theme(legend.position = "bottom")

p1 + p2
```

Running the `core_periphery()` function on this idealized graph should give the optimal result.
```{r}
#| label: split-cp
core_periphery(sg)
```

The function returns a list with two entries. The first, `vec`, returns the membership of nodes. It is one if a node is in the core and 0 if not. The second entry `corr` is the correlation of the adjacency matrix with the ideal adjacency matrix given from the `vec` memberships. In this case, the correlation is one meaning that we have discovered the perfect core-periphery structure. In empirical applications, we rarely expect such a good fit. 

The function also has an argument which allows to use different optimization techniques to find core-periphery structures (the problem of finding the optimal solution is too complex). For illustration, we rewire some edges of the  graph `sg` to destroy the idealized structure and use all implemented optimization methods.

```{r}
#| label: rewired-cp
set.seed(45)
sg1 <- rewire(sg, each_edge(0.25))
core_periphery(sg1, method = "rk1_dc")
core_periphery(sg1, method = "rk1_ec")
core_periphery(sg1, method = "GA")
```

`rk1_dc` and `rk1_ec` are so called rank-one matrix approximation methods which infer an idealized structure via the degrees and eigenvectors of the adjacency matrix. These methods are extremely fast but might result in a lower quality of the result compared to `GA` which runs a genetic algorithm. 

There are several extensions to this simple discrete core-periphery model, including versions for weighted and directed networks. An extension for weighted networks, designed for international trade data, is implemented in the package `ITNr`.

## References {.unnumbered}

Clauset, A.; Newman, M. E. J. & Moore, C. Finding community structure in very large networks, Physical Review E 2004, 70, 066111

Vincent D. Blondel, Jean-Loup Guillaume, Renaud Lambiotte, Etienne Lefebvre: Fast unfolding of communities in large networks. J. Stat. Mech. (2008) P10008

Traag, V. A., Waltman, L., & van Eck, N. J. (2019). From Louvain to Leiden: guaranteeing well-connected communities. Scientific reports, 9(1), 5233.

Žiberna, A. (2007). Generalized Blockmodeling of Valued Networks. Social Networks, 29(1), 105-126. doi: 10.1016/j.socnet.2006.04.002

Žiberna, A. (2008). Direct and indirect approaches to blockmodeling of valued networks in terms of regular equivalence. Journal of Mathematical Sociology, 32(1), 57-84. doi: 10.1080/00222500701790207

Žiberna, A. (2014). Blockmodeling of multilevel networks. Social Networks, 39(1), 46-61. doi: 10.1016/j.socnet.2014.04.002

Borgatti, Stephen P., and Martin G. Everett. "Models of core/periphery structures." Social networks 21.4 (2000): 375-395.
