# Ego Networks
Ego networks refer to the personal network surrounding a single focal individual, known as the *ego*. An ego network includes the ego, the people directly connected to them, so called *alters*, and the relationships among those alters. Rather than examining an entire community or organization, ego network analysis focuses on the immediate social environment of one person, allowing researchers to understand how information, resources, influence, or support flow through that individual’s connections. By analyzing the size of the network, the strength of ties, the diversity of contacts, and the level of interconnectedness among alters, scholars can assess factors such as social capital, access to opportunities, and resilience. Ego networks are especially useful in studying phenomena like job searching, knowledge sharing, health behaviors, and online social interaction, where an individual’s direct relationships play a critical role in shaping outcomes.


In this chapter, we will begin by introducing the fundamentals of ego network data and learning how to use the `egor` package to manipulate, construct, and visualize ego networks. We will then examine substantive research questions related to homophily; the tendency for similar individuals to associate with one another more frequently than with dissimilar others, across various demographic dimensions. Finally, we will explore an applied example in which properties of ego networks are used to predict important outcomes of interest.

## Packages Needed for this Chapter
```{r}
#| message: false
library(igraph)
library(egor)
library(networkdata)
library(ggraph)
library(patchwork)
library(tidyverse)
```

::: {.callout-note}
## Note

The `egor` package works with ego network data in *tidy* format, meaning it follows tidy data principles. In tidy data, each row represents one observation and each column represents one variable. Instead of storing everything in a single large matrix, ego network data are separated into three linked tables: one containing information about egos, one containing information about alters, and one containing the ties among alters. These tables are connected through ID variables that indicate which alters belong to which ego and which alters are tied to one another. This structure makes the data easier to understand, manipulate, and analyze using tools from the tidyverse, while still preserving the relational nature of network data.
:::

## Ego Network Data Structure

Ego network data are typically generated through survey-based approaches that capture information about individuals and their immediate social surroundings. In one common design, respondents (*egos*) are asked to name the people with whom they have specific types of relationships and to describe the connections among those named individuals (*alters*), thereby providing both node-level and tie-level information from a single informant. A more elaborate strategy uses a two-stage snowball approach: after the initial respondent identifies their contacts, those contacts are subsequently surveyed about their relationships with one another, allowing for more detailed mapping of ties within the ego’s local network. Although data collected in these ways do not reveal how these personal networks are embedded within the broader population structure, they enable researchers to estimate the distribution and characteristics of ego network forms across even very large populations.

An important methodological implication of this design is that survey-based ego network data typically produce independent observations: each ego network is centered on a different respondent and represents a separate local structure. As a result, ego networks can often be treated as independent units of analysis, which allows researchers to apply standard statistical models (such as regression models) to explain variation in network characteristics (e.g., density, composition, or diversity) across egos.

Methodologically, such data consist of multiple distinct networks, each centered on a different ego and typically composed of different actors. This means that each ego network should be stored and analyzed as a separate actor-by-actor matrix rather than combined into a single stacked dataset, since the matrices do not represent multiple relations among the same set of individuals.

A second common source of ego network data involves deriving them from complete network datasets that already contain information about all actors and ties within a bounded population. In this case, researchers can extract the local network surrounding a focal actor directly from the larger structure. Simple extraction procedures allow analysts to isolate an ego and their direct ties, though these methods may omit information about relationships among the alters themselves. More comprehensive approaches rely on identifying a defined neighborhood—such as through a previously constructed attribute or partition that marks members of a particular ego’s local environment—and then generating a subgraph that preserves both ego–alter and alter–alter ties. Unlike survey-generated ego networks, these extracted ego networks remain embedded within a known global structure, enabling comparisons between local and whole-network properties.


### Constructing an Ego Network with `egor`
An `egor` object stores the multiple data levels required for ego-centered network analysis. These levels include **ego-level data** (attributes of respondents), **alter-level data** (attributes of the individuals named by the respondent), and **alter–alter ties** (relationships among those alters).

The data used in the following examples come from the 2004 General Social Survey (GSS) egocentric network module, which is available in the `networkdata` package. The dataset contains a stratified sample of 594 respondents and provides information at all three levels commonly used in ego network analysis: characteristics of the respondents (*egos*), characteristics of the individuals they name in their networks (*alters*), and the relationships among those alters (*alter–alter ties*). In the package, these data are already provided in the form of an `egor` object.

To explore the structure of the dataset, we can inspect the first few rows of each component of the `egor` object:

```{r}
#| label: load-ego-data

data("gss_egor")
head(gss_egor$ego) # ego level
head(gss_egor$alter) # alter level
head(gss_egor$aatie) # alter-alter ties
```

Each of the three data frames is linked through unique identifier variables that ensure the different levels of information correspond to the correct individuals. The `.egoID` variable appears in all three tables and serves as the key that connects each respondent’s attributes to their alters and to the ties among those alters. In the alter-level data, each network member receives a unique `.altID`. These identifiers are referenced again in the alter–alter tie table through the `.srcID` (source) and `.tgtID` (target) columns, which specify which two alters are connected. The `weight` column records the value or strength of the tie between them.

Before proceeding with the analysis, we remove respondents for whom the variable `numgiven` is missing in the ego-level data. This variable records how many alters each respondent named and is required for several of the descriptive measures used later in the chapter.  Because `numgiven` records how many alters a respondent named, a missing value at the ego level implies that no alter information is recorded for that respondent. Consequently, it is sufficient to remove missing values only from the ego table, since the alter and alter–alter tie tables contain entries only for egos who actually reported network members.


```{r}
#| label: na-remove
# Remove egos with missing numgiven
gss_egor$ego <- gss_egor$ego[!is.na(gss_egor$ego$numgiven), ]
```

Examining the dimensions of the three components after removing missing values (`dim(gss_egor$ego)`, `dim(gss_egor$alter)`, and `dim(gss_egor$aatie`) provides a quick overview of the dataset. The ego table now contains 305 respondents. Across these respondents, a total of 555 alters were named, each linked to a specific ego via `.egoID` and uniquely identified within each ego network by `.altID`. The alter–alter tie table contains 640 rows, with each row representing a relationship between two alters within the same ego network. These ties are identified by the `.egoID` of the corresponding ego and the `.srcID` and `.tgtID` variables that indicate the two alters involved.

## Plotting Ego Nets

We now plot the networks using `igraph`. The first step is to convert the `egor` object into a list of `igraph` objects using `as_igraph()` from the `egor` package. Below, we create the list and inspect the first three ego networks.

```{r}
#| label: egor-2-igraph
# Convert egor object to a list of igraph ego networks
gss_g <- as.igraph(gss_egor)

# Inspect the first three ego networks
gss_g[1:3]
```

::: {.callout-note}
If we instead wanted to work within the `statnet` framework, we could use `as_network()` to convert the data into objects of class `network`.
:::

After running `as_igraph()`, we obtain a list of ego networks in igraph format, where each element of the list corresponds to one ego's personal network. The attributes stored at the alter level are automatically transferred as vertex attributes, and the alter–alter tie information (including edge weights) is preserved as edge attributes. By default, the ego node itself is not included in these igraph representations. This is common practice because ego is, by definition, connected to all alters, and therefore contributes little additional structural variation in many visualizations or structural measures. In later sections, we will examine statistics that explicitly incorporate both ego and alter information.

We now visualize a subset of the data by plotting the first 9 ego networks in the list using `ggraph`. This is shown in @fig-egos.

::: {#fig-egos}
```{r}
#| label: plot-egos
#| message: false
#| fig-height: 9
#| fig-width: 9

plots <- lapply(gss_g[1:9], function(g) {
  
  # rebuild a clean graph using only structure + names
  el <- as_edgelist(g, names = TRUE)
  vn <- V(g)$name
  
  if (length(el) == 0 || nrow(el) == 0) {
    g_clean <- make_empty_graph(n = length(vn), directed = FALSE)
    V(g_clean)$name <- vn
  } else {
    g_clean <- graph_from_edgelist(el, directed = FALSE)
    
    # add back isolated vertices, if any
    missing_v <- setdiff(vn, V(g_clean)$name)
    if (length(missing_v) > 0) {
      g_clean <- add_vertices(g_clean, nv = length(missing_v), name = missing_v)
    }
  }
  
  ggraph(g_clean, layout = "stress") +
    geom_edge_link(alpha = 0.4) +
    geom_node_point(size = 6, color = "steelblue") +
    geom_node_text(aes(label = name), size = 3, color = "white") +
    theme_graph()
})

wrap_plots(plots, ncol = 3)
```
The first nine ego networks derived from the egor object based on the example dataset. Each panel represents the alter–alter structure of a single ego’s personal network, where nodes correspond to alters and edges represent ties among them. The ego node itself is not shown, as it is by definition connected to all alters. Differences across panels illustrate variation in density, clustering, and overall structural configuration across ego networks.
:::

## Descriptive Statistics for Ego Networks
One of the most commonly used structural measures in ego network analysis is **density**. Density captures the proportion of all possible alter–alter ties that are actually present in an ego network. In other words, it tells us how interconnected an ego’s alters are with one another.  For example, if an ego has 5 alters, there are
$$
\frac{5 \times (5 - 1)}{2} = 10
$$
possible undirected ties among those alters. If all 10 ties are present, density equals 1 (a fully connected clique). If only 8 of the 10 possible ties are present, density equals 0.8.


Density is often of substantive interest because it reflects the level of cohesion within a personal network. Highly dense ego networks (values close to 1) indicate tightly knit groups in which most alters are connected to one another. Such cohesive structures can facilitate trust, norm enforcement, and social support, as information circulates quickly and reputational mechanisms are strong. However, according to **structural hole theory** [@burt1992structural],  dense networks tend to generate redundant information because alters are connected to many of the same people and share overlapping perspectives. The key advantages of social capital derive not primarily from the attributes of one’s alters, but from the *structure* of the ego network itself. Individuals benefit when they bridge gaps, so-called structural holes, between otherwise disconnected groups. In low-density networks, ego may occupy a brokerage position linking clusters that are not directly connected to each other, thereby gaining access to diverse, non-redundant information and potential strategic advantages (such as faster promotion within organizations).  Density therefore captures an important theoretical trade-off between cohesion and brokerage in ego network structure.

We can calculate density for each ego network using the `ego_density()` function applied to our `ego32` object. Note that ego networks with 0 or 1 alters will have `NA` values for density, because density is undefined when no alter–alter ties are possible.

```{r}
#| label: ego-dens

dens <- ego_density(gss_egor)
head(dens)
```
