# Entropy Analysis  {#sec-netropy}
A central goal in the analysis of network data is to understand the dependence structure among variables describing entities and their interactions. Traditional statistical methods typically focus on pairwise relationships and often rely on linear assumptions. However, network data frequently exhibit complex, nonlinear, and higher-order dependencies that are not adequately captured by such approaches.

Multivariate entropy analysis provides a flexible and general framework for addressing these challenges. By quantifying uncertainty and information in multivariate distributions, entropy-based methods allow for the detection and characterization of dependence structures without imposing restrictive parametric assumptions. In particular, they are well suited for identifying interactions among multiple variables simultaneously, including those defined on different components of a network.

To apply entropy-based methods effectively, variables must be represented on finite discrete range spaces and aligned with a common domain of observation. This often requires preprocessing steps such as discretization of continuous variables and careful consideration of how observations are defined, for example as nodes, dyads, or temporal snapshots. These steps are crucial for reliable estimation of entropy measures and meaningful interpretation of dependence patterns.

In this chapter, we focus on the application of multivariate entropy analysis to network data, with particular attention to how data representation, discretization, and domain structure influence the identification of dependencies.


## Packages Needed for this Chapter
```{r}
#| message: false
#| warning: false
library(netropy)
library(igraph)
library(ggplot2)
```

## Network Data Representation and Transformation
Network data consist of variables defined on different domains, most commonly nodes (vertices) and dyads (edges). While vertex variables are observed on $n$ nodes, edge variables are observed on $\binom{n}{2}$ dyads. Since entropy analysis requires all variables to be defined on a common domain, these differences must be resolved through appropriate transformations.

Let $V = {1, \dots, n}$ denote a set of nodes. Suppose $X_i$, $i = 1, \dots, m’$, are vertex variables with $r_i$ outcomes, and $Y_j$, $j = 1, \dots, m’’$, are edge variables with $r’_j$ outcomes. The corresponding joint outcome sizes are $r’ = \prod r_i$ and $r’’ = \prod r’_j$. Because these variables are observed on different domains, they cannot be directly combined in entropy calculations.

One possibility is to transform vertex variables to the edge domain. For each pair $(u,v)$, define $X_i^* = (X_u, X_v)$. This yields a dataset with $\binom{n}{2}$ observations, allowing the transformed vertex variables to be analyzed jointly with edge variables.

#### Table: Edge-Based Representation {.unnumbered}

| observation | $X_1^*$ | $X_2^*$ | $\cdots$ | $X_{m'}^*$ | $Y_1$ | $Y_2$ | $\cdots$ | $Y_{m''}$ |
|------------|--------|--------|----------|------------|------|------|----------|------------|
| 1          |        |        |          |            |      |      |          |            |
| 2          |        |        |          |            |      |      |          |            |
| $\vdots$   |        |        |          |            |      |      |          |            |
| $\binom{n}{2}$ |    |        |          |            |      |      |          |            |


An alternative is to transform edge variables to the vertex domain. For each node $v$, define $Y_j^* = \sum_u Y_{uv}$. This produces vertex-level summaries such as degrees or centrality-type measures and results in a dataset with $n$ observations.


#### Table: Vertex-Based Representation  {.unnumbered}

| observation | $X_1$ | $X_2$ | $\cdots$ | $X_{m'}$ | $Y_1^*$ | $Y_2^*$ | $\cdots$ | $Y_{m''}^*$ |
|------------|------|------|----------|----------|--------|--------|----------|------------|
| 1          |      |      |          |          |        |        |          |            |
| 2          |      |      |          |          |        |        |          |            |
| $\vdots$   |      |      |          |          |        |        |          |            |
| $n$        |      |      |          |          |        |        |          |            |


A useful extension is to construct edge attributes directly from vertex attributes. For example, if $(W_i, T_i, P_i)$ are binary vertex variables for node $i$, then edge-level variables can be defined by $W_{ij} = W_i + W_j$, $T_{ij} = T_i + T_j$, and $P_{ij} = P_i + P_j$ [@frank2016multivariate]. These new variables take values in ${0,1,2}$ and can be analyzed jointly with other dyadic variables.

These transformations involve trade-offs. Extending vertex variables to the edge domain increases the number of observations but also enlarges the joint outcome space. Aggregating edge variables to the vertex domain simplifies the data structure but may reduce variability and finer relational detail. Nevertheless, these transformations are essential because they make it possible to bring variables from different domains into a common analytical framework. The same principles extend naturally to higher-order structures such as triads.

Once variables have been aligned to a common domain and appropriately discretized, the data are ready for multivariate entropy analysis. Entropy can be used for dimensionality reduction by identifying redundant variables and selecting informative subsets. Mutual information, derived from joint entropy, quantifies the strength and form of dependence between variables, including higher-order interactions. Expected conditional entropies can be used to assess predictive power by measuring how much uncertainty about one variable is reduced given knowledge of others. Together, these measures provide a comprehensive approach to understanding association, dependence, and information flow in network data. In the following sections, we illustrate this framework through an application.

## Network Study of Corporate Law firm 
The first application considers network data from a corporate law firm in the Northeastern U.S.[@lazega2001collegial] which is available in the `netropy` package. 

The network consists of 71 lawyers connected through three types of informal interactions: co-working (undirected), advice-seeking (directed), and friendship (directed). In addition, each lawyer is described by several attributes, including status (associate or partner), gender, office location (Boston, Hartford, or Providence), years at the firm, age, and area of practice (corporate or litigation).


To load the data, we extract the adjacency matrices corresponding to the three network relations (advice, friendship, and co-working) as well as the dataframe containing node-level attributes for each lawyer. The objects are stored in a list and can be assigned as follows:

```{r}
#| label: load-data
data(lawdata) 
adj_advice <- lawdata[[1]]   # adjacency matrix: advice network (directed)
adj_friend <- lawdata[[2]]   # adjacency matrix: friendship network (directed)
adj_cowork <- lawdata[[3]]   # adjacency matrix: co-working network (undirected)
df_att <- lawdata[[4]]       # node attributes for the 71 lawyers
```


In the following analyses we also consider three additional networks created from the observed ones above. We create new dyad variables for mutual friendship and mutual advice, and likely friendships which is an undirected network and does not require reciprocal reporting of friendships. The latter is done by creating the directed graph object and then extracting the adjacency matrix from the graph assuming it is undirected instead:

```{r}
#| label: new-vars

# mutual advice
adj_mutadv <- adj_advice*t(adj_advice)
# mutual friendship
adj_mutfrn <- adj_friend*t(adj_friend) 
# likely friendship 
g_frn <- graph_from_adjacency_matrix(adj_friend, mode="directed") 
g_likfrn <- as_undirected(g_frn)
adj_likfrn <-  as_adjacency_matrix(g_likfrn, sparse = FALSE)
```



### Data Editing
#### Vertex Variables {.unnumbered}

Observed node variables are the attributes in the data frame `df_att`. The first 5 lines of this data frame are:
```{r}
head(df_att, 5)
```
All variables, except `years` and `age`, are categorical with finite range spaces and therefore kept in their original form. 
To incorporate the continuous variables `years` and `age` into the entropy analysis, we first transform them into ordinal variables with a small number of categories. Since entropy estimation requires finite range spaces with sufficiently many observations in each category, we use the empirical cumulative distribution functions (CDFs) of these variables to identify cutpoints that produce approximately equally sized groups.

We begin by constructing frequency tables for `years` and `age`, and then compute the corresponding relative and cumulative relative frequencies.
```{r}
#| label: cdf
# Years CDF data
frq_years <- table(df_att$years) |>
  as.data.frame()

names(frq_years) <- c("years", "freq")
frq_years$years <- as.numeric(as.character(frq_years$years))
frq_years$rel.freq <- frq_years$freq / sum(frq_years$freq)
frq_years$cum.rel.freq <- cumsum(frq_years$rel.freq)

# Age CDF data
frq_age <- table(df_att$age) |>
  as.data.frame()

names(frq_age) <- c("age", "freq")
frq_age$age <- as.numeric(as.character(frq_age$age))
frq_age$rel.freq <- frq_age$freq / sum(frq_age$freq)
frq_age$cum.rel.freq <- cumsum(frq_age$rel.freq)
```

The empirical CDFs make it possible to inspect how the observations accumulate across the support of each variable. This is shown in  @fig-cdfs.
We use these plots to choose cutpoints that divide the data into three approximately equally sized groups. For years, the selected cutpoints are $3$ and $13$, while for age, the selected cutpoints are $35$ and $45$.

::: {#fig-cdfs}
```{r}
#| label: plot-years-cdf-gg
#| echo: false
#| fig-height: 5
#| fig-width: 10
library(patchwork)
pyears <- ggplot(frq_years, aes(x = years, y = cum.rel.freq)) +
  geom_step(linewidth = 1) +
  geom_point() +
  geom_hline(yintercept = c(1/3, 2/3), linetype = "dashed") +
  geom_vline(xintercept = c(3, 13), linetype = "dotted") +
  labs(
    title = "CDF of Years at Firm",
    x = "Years at firm",
    y = "Cumulative relative frequency"
  ) +
  theme_minimal()

page <- ggplot(frq_age, aes(x = age, y = cum.rel.freq)) +
  geom_step(linewidth = 1) +
  geom_point() +
  geom_hline(yintercept = c(1/3, 2/3), linetype = "dashed") +
  geom_vline(xintercept = c(35, 45), linetype = "dotted") +
  labs(
    title = "CDF of Age",
    x = "Age",
    y = "Cumulative relative frequency"
  ) +
  theme_minimal()

pyears + page
```
Empirical cumulative distribution functions (CDFs) for `years at firm` and `age`. The dashed horizontal lines indicate the target quantiles (approximately one-third and two-thirds), while the dotted vertical lines show the selected cutpoints used to define three approximately equally sized categories for each variable.
:::


Using these cutpoints, we create a new dataframe `att_var` that contains the variables used in the analysis. All variables except years and age are already categorical with finite range spaces and are therefore retained in their original form, apart from simple recoding to start at $0$ where needed. The variables years and age are recoded into three ordinal categories.


By looking at these cdf's we can find values with which we base the categories on:
```{r}
#| label: categorized-vars
att_var <- data.frame(
  status    = df_att$status - 1,
  gender    = df_att$gender,
  office    = df_att$office - 1,
  years     = ifelse(df_att$years <= 3, 0,
                ifelse(df_att$years <= 13, 1, 2)),
  age       = ifelse(df_att$age <= 35, 0,
                ifelse(df_att$age <= 45, 1, 2)),
  practice  = df_att$practice,
  lawschool = df_att$lawschool - 1
)
```

Finally, we examine the resulting categorized distributions for years and age to confirm that the selected cutpoints produce reasonably balanced groups. This helps ensure that the discretized variables are suitable for subsequent entropy-based analysis. This is shown in @fig-cats.


::: {#fig-cats}

```{r}
#| label: categorized-data-base
#| echo: false
#| fig-height: 5
#| fig-width: 10
years_cat <- as.data.frame(table(att_var$years))
names(years_cat) <- c("category", "freq")
years_cat$variable <- "years"

age_cat <- as.data.frame(table(att_var$age))
names(age_cat) <- c("category", "freq")
age_cat$variable <- "age"

att_long <- rbind(years_cat, age_cat)

ggplot(att_long, aes(x = factor(category), y = freq)) +
  geom_col(fill = "lightgrey", color = "darkgrey") +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    x = "Category",
    y = "Frequency"
  ) +
  theme_minimal()
```
Frequency distributions of the discretized variables `age` and `years`, showing approximately balanced categories after CDF-based grouping.
:::

This preprocessing step yields a set of discrete variables with manageable range sizes, making them appropriate for estimating joint and conditional entropies in the subsequent network analysis.

Next, we transform the observed dyad variables into node-level variables. This is achieved by aggregating edge information into node-based measures. In particular, we use degree-based summaries: in-degree and out-degree for the directed advice and friendship networks, and degree (number of ties) for the undirected co-work network, as well as for the mutual advice, mutual friendship, and likely friendship networks.

To compute these measures, we first construct graph objects from each adjacency matrix (note that the directed friendship graph `g_frn` and the undirected likely-friendship graph `g_likfrn` were already created above):

```{r}
#| label: g-objects
g_cwk    <- graph_from_adjacency_matrix(adj_cowork, mode = "undirected")
g_adv    <- graph_from_adjacency_matrix(adj_advice, mode = "directed")
g_mutadv <- graph_from_adjacency_matrix(adj_mutadv, mode = "undirected")
g_mutfrn <- graph_from_adjacency_matrix(adj_mutfrn, mode = "undirected")
```

Degrees are then computed as

```{r}
#| label: degs
deg_cwk     <- degree(g_cwk)
indeg_adv   <- degree(g_adv, mode = "in")
outdeg_adv  <- degree(g_adv, mode = "out")
deg_mutadv  <- degree(g_mutadv)
indeg_frn   <- degree(g_frn, mode = "in")
outdeg_frn  <- degree(g_frn, mode = "out")
deg_mutfrn  <- degree(g_mutfrn)
deg_likfrn  <- degree(g_likfrn)
```
These variables are then categorized using the same procedure as above for the other vertex variables:
```{r}
#| label: degs-cats
deg_var <- data.frame(
  deg_cwk     = ifelse(deg_cwk <= 9, 0, 1),
  indeg_adv   = ifelse(indeg_adv <= 10, 0, 1),
  outdeg_adv  = ifelse(outdeg_adv <= 11, 0, 1),
  deg_mutadv  = ifelse(deg_mutadv <= 3, 0, 1),
  indeg_frn   = ifelse(indeg_frn <= 7, 0, 1),
  outdeg_frn  = ifelse(outdeg_frn <= 7, 0, 1),
  deg_mutfrn  = ifelse(deg_mutfrn <= 4, 0, 1),
  deg_likfrn  = ifelse(deg_likfrn <= 11, 0, 1)
)
```
Merging `deg_var` and `att_var` yields the final data frame `vertex_var`, which contains all 15 observed and derived vertex variables used in the subsequent entropy analysis:

```{r}
#| label: vertex-vars
vertex_var <- cbind(att_var, deg_var)
```
The resulting dataset `vertex_var` contains all discretized vertex-level variables used in the entropy analysis. The first rows are shown in @tbl-vertex-var.


::: {#tbl-vertex-var}
```{r}
#| echo: false
knitr::kable(
 head(vertex_var, 10),
)
```
First 10 rows of the vertex-level dataset.
:::


#### Dyad Variables {.unnumbered}

Dyad variables are constructed as pairs of incident vertex variables, resulting in $\binom{71}{2} = 2485$ observations (i.e., the number of rows in the dyadic data frames created below). 

The observed node attributes in the data frame `att_var` are therefore transformed into dyadic attributes by pairing the values of two connected vertices. For example, the binary variable `status` yields four possible dyadic outcomes: $(0,0)$, $(0,1)$, $(1,0)$, and $(1,1)$. Similarly, the categorical variable `office` with three levels results in nine possible dyadic outcomes: $(0,0)$, $(0,1)$, $(0,2)$, $(1,0)$, $(1,1)$, $(1,2)$, $(2,0)$, $(2,1)$, and $(2,2)$.

These transformations can be carried out using the function `get_dyad_var()`, where the argument `type = "att"` specifies that the input consists of vertex attributes. This procedure is applied to all vertex variables in the dataset:

```{r}
#| label: dyad-transf
dyad_status    <- get_dyad_var(att_var$status, type = "att")
dyad_gender    <- get_dyad_var(att_var$gender, type = "att")
dyad_office    <- get_dyad_var(att_var$office, type = "att")
dyad_years     <- get_dyad_var(att_var$years, type = "att")
dyad_age       <- get_dyad_var(att_var$age, type = "att")
dyad_practice  <- get_dyad_var(att_var$practice, type = "att")
dyad_lawschool <- get_dyad_var(att_var$lawschool, type = "att")
```

Note that the dyadic outcomes are recoded to numerical values to avoid character objects when performing the entropy analysis. In practice, the actual labels of the categories are irrelevant, as entropy depends only on the frequencies of occurrence. For example, the variable `status` takes values $0$–$3$, while `office` takes values $0$–$8$.

In a similar way, we construct dyad variables based on observed and derived ties. For undirected networks, dyad variables are simple indicator variables obtained directly from the adjacency matrix. For directed networks, each dyad is represented by a pair of indicators corresponding to sending and receiving ties, resulting in four possible outcomes (e.g., for `advice` and `friendship`).

Using the same function as above, we now set the argument `type = "tie"` and provide the adjacency matrix as input. The function automatically determines whether the network is directed or undirected and creates the appropriate dyadic variable.

```{r}
#| label: dyad-transf2
dyad_cwk    <- get_dyad_var(adj_cowork, type = "tie")
dyad_adv    <- get_dyad_var(adj_advice, type = "tie")
dyad_mutadv <- get_dyad_var(adj_mutadv, type = "tie")
dyad_frn    <- get_dyad_var(adj_friend, type = "tie")
dyad_mutfrn <- get_dyad_var(adj_mutfrn, type = "tie")
dyad_likfrn <- get_dyad_var(adj_likfrn, type = "tie")
```




Next, all dyad variables are merged into a single data frame for subsequent entropy analysis:
```{r}
#| label: dyad-var
dyad_var <- data.frame(
  status    = dyad_status$var,
  gender    = dyad_gender$var,
  office    = dyad_office$var,
  years     = dyad_years$var,
  age       = dyad_age$var,
  practice  = dyad_practice$var,
  lawschool = dyad_lawschool$var,
  cowork    = dyad_cwk$var,
  advice    = dyad_adv$var,
  mutadv    = dyad_mutadv$var,
  friend    = dyad_frn$var,
  mutfrn    = dyad_mutfrn$var,
  likfrn    = dyad_likfrn$var
)
```
The resulting dataset `dyad_var` contains all dyadic variables used in the entropy analysis. The first rows are shown in @tbl-dyad-var.

::: {#tbl-dyad-var}
```{r}
#| echo: false
knitr::kable(
  head(dyad_var, 10)
)
```
First 10 rows of the dyadic dataset.
:::

#### Triad Variables {.unnumbered}
A similar function, `get_triad_var()`, is used to transform vertex variables and different relation types into triad-level variables. These variables represent combinations of three nodes and therefore yield $\binom{71}{3} = 57155$ observations in the law firm dataset. Each observation corresponds either to a triple of individual attributes or to the configuration of relations among three nodes.

As for dyad variables, the function requires specifying the type of input: `type = "att"` for vertex attributes (given as column vectors) and `type = "tie"` for relational data (given as adjacency matrices).

For vertex variables, triad variables are constructed as follows:

```{r}
#| label: triad-var
triad_status    <- get_triad_var(att_var$status, type = "att")
triad_gender    <- get_triad_var(att_var$gender, type = "att")
triad_office    <- get_triad_var(att_var$office, type = "att")
triad_years     <- get_triad_var(att_var$years, type = "att")
triad_age       <- get_triad_var(att_var$age, type = "att")
triad_practice  <- get_triad_var(att_var$practice, type = "att")
triad_lawschool <- get_triad_var(att_var$lawschool, type = "att")
```

Binary vertex attributes result in $2^3 = 8$ possible triadic outcomes:
$$
(0,0,0), (1,0,0), (0,1,0), (1,1,0), (0,0,1), (1,0,1), (0,1,1), (1,1,1),
$$
which are recoded as integers from $0$ to $7$. Similarly, attributes with three categories yield $3^3 = 27$ possible triadic outcomes, coded from $0$ to $26$.

For relational variables, undirected ties are transformed from binary outcomes into triad variables with $8$ possible configurations, while directed ties are transformed from $4$ dyadic outcomes into $4^3 = 64$ triadic configurations, capturing all possible combinations of sending and receiving ties among three nodes.

The triad variables based on ties are constructed as follows:
```{r}
#| label: triad-var2
triad_cwk    <- get_triad_var(adj_cowork, type = "tie")
triad_adv    <- get_triad_var(adj_advice, type = "tie")
triad_mutadv <- get_triad_var(adj_mutadv, type = "tie")
triad_frn    <- get_triad_var(adj_friend, type = "tie")
triad_mutfrn <- get_triad_var(adj_mutfrn, type = "tie")
triad_likfrn <- get_triad_var(adj_likfrn, type = "tie")
```
All triad variables are then merged into a single data frame for subsequent entropy analysis:
```{r}
#| label: triad-var-df
triad_var <- data.frame(
  status    = triad_status$var,
  gender    = triad_gender$var,
  office    = triad_office$var,
  years     = triad_years$var,
  age       = triad_age$var,
  practice  = triad_practice$var,
  lawschool = triad_lawschool$var,
  cowork    = triad_cwk$var,
  advice    = triad_adv$var,
  mutadv    = triad_mutadv$var,
  friend    = triad_frn$var,
  mutfrn    = triad_mutfrn$var,
  likfrn    = triad_likfrn$var
)
```

The resulting dataset `triad_var` contains all triad-level variables used in the entropy analysis. The first rows are shown in @tbl-triad-var.

::: {#tbl-triad-var}
```{r}
#| echo: false
knitr::kable(
  head(triad_var, 10)
)
```
First 10 rows of the triad-level dataset.
:::



### Univariate and Bivariate Entropies
The univariate entropy of a discrete variable $X$ with $r$ outcomes is defined as
$$
H(X) = \sum_x p(x) \log_2 \frac{1}{p(x)}.
$$
It measures the uncertainty of $X$. A variable with zero entropy has no uncertainty (it is constant) and can be omitted from further analysis. The maximum entropy is $\log_2 r$, which corresponds to a uniform distribution over all outcomes.

The bivariate entropy of two discrete variables $X$ and $Y$ is given by
$$
H(X,Y) = \sum_x \sum_y p(x,y) \log_2 \frac{1}{p(x,y)}.
$$
This quantity can be used to assess redundancy, functional relationships, and stochastic independence. It satisfies the bounds
$$
H(X) \leq H(X,Y) \leq H(X) + H(Y).
$$

- The lower bound is attained if and only if there is a functional relationship $Y = f(X)$.
- The upper bound is attained if and only if $X$ and $Y$ are stochastically independent, $X \bot Y$.

In particular, if $H(X,Y) = H(X)$ or $H(X,Y) = H(Y)$, one of the variables is redundant and does not provide additional information.

The function `entropy_bivar()` computes the bivariate entropies for all pairs of variables in a data frame. The results are returned as an upper triangular matrix, where each cell contains the joint entropy of the corresponding row and column variables. The diagonal entries represent the univariate entropies of each variable.

#### Vertex Variables {.unnumbered}

We apply the function to the data frame containing all observed and transformed vertex variables.

```{r}
#| label: vert-ent
h2_vertex <- entropy_bivar(vertex_var)
```

The output, rounded to 2 decimals, is shown in @tbl-ent-vertvar.

::: {#tbl-ent-vertvar}
```{r}
#| echo: false
knitr::kable(
  round(h2_vertex, 2),
)
```
Bivariate entropy matrix for vertex variables.
:::

Bivariate entropies can be used to detect redundant variables that do not contribute additional information and may therefore be omitted from further analysis. Redundancy can be identified by comparing diagonal entries (univariate entropies) with the corresponding off-diagonal values in the same rows or columns. If these values are equal, it indicates that one variable does not provide additional information beyond another.

This can also be assessed using the function `redundancy()`. The output of `redundancy()` is a binary matrix indicating whether one variable is redundant with respect to another. A value of `1` means that the corresponding column variable does not provide additional information beyond the row variable, while `0` indicates that the variables contain distinct information.

For the vertex variables in this dataset, no redundancy is detected:
```{r}
#| label: ver-red
red_vertex <- redundancy(vertex_var, dec = 3)
```
Note that the argument `dec` affects the detection of redundant variables, as it controls the rounding of bivariate entropies. Using fewer decimal places increases the likelihood that two variables will appear to have identical entropy values and thus be flagged as redundant.

For illustration, we apply the `redundancy()` function to the original attribute data frame `df_att`. Recall that the variable `senior` consists entirely of unique values and is therefore identified as redundant, as shown below:

```{r}
#| label: ver-red2
red_vertex2 <- redundancy(df_att, dec = 2)
red_vertex2
# find columns that are redundant with any other variable
colnames(red_vertex2)[rowSums(red_vertex2) >= 1]

```

Here, the variable `senior` is flagged as redundant with respect to all other variables, as indicated by the row of ones. This occurs because `senior` consists entirely of unique values, meaning it does not share any repeated structure with other variables and therefore does not contribute meaningful information for entropy-based analysis. In contrast, all other variables show zeros, indicating no redundancy among them.

#### Dyad Variables {.unnumbered}
The data frame `dyad_var` contains 13 observed and transformed dyad-level variables. To compute the univariate and bivariate entropies, we apply the function `entropy_bivar()`:

```{r}
#| label: dyad-ent
h2_dyad <- entropy_bivar(dyad_var)
```
The results shown in @tbl-ent-dyadvar indicate redundancy among the different variants of the advice and friendship relations. This is expected, as these variables are derived from the same underlying ties. 

::: {#tbl-ent-dyadvar}
```{r}
#| echo: false
knitr::kable(
  round(h2_dyad, 2),
)
```
Bivariate entropy matrix for dyad variables. Redundancy is observed among the advice and friendship variants: the variables `mutadv`, `mutfrn`, and `likfrn` do not provide additional information beyond `advice` and `friend`, respectively, and should therefore excluded from subsequent analyses.
:::


The redundancy can also be confirmed using the `redundancy()` function. Based on this analysis, we remove the redundant variables mutadv, mutfrn, and likfrn (columns 10, 12, and 13) from the data frame:
```{r}
dyad_var <- dyad_var[-c(10, 12:13)]
```

The resulting data frame contains the non-redundant dyad variables used in the subsequent entropy analysis.

#### Triad Variables {.unnumbered}


The data frame `triad_var` contains the observed and transformed triad-level variables. To compute the univariate and bivariate entropies, we apply the function `entropy_bivar()`:

```{r}
#| label: triad-ent
h2_triad <- entropy_bivar(triad_var)
```
The results shown in @tbl-ent-triadvar reveal redundancy among the triad variables derived from the different variants of the advice and friendship relations. As in the dyadic case, this is expected since these variables originate from the same underlying ties.

::: {#tbl-ent-triadvar}

```{r}
#| echo: false
knitr::kable(
  round(h2_triad, 2),
)
```

Bivariate entropy matrix for triad variables.  Redundancy is observed among the triad variables derived from the different variants of the advice and friendship relations.
:::

The redundancy can be confirmed using the `redundancy()` function and after removing the redundant variables (`mutadv`, `mutfrn`, and `likfrn`), the resulting data frame contains 10 triad variables, which are used in the subsequent entropy analysis.

```{r}
triad_var <- triad_var[-c(10, 12:13)]
```

### Joint Entropies and Association Graphs
Joint entropy measures the total uncertainty associated with a pair of variables $(X,Y)$ and provides a basis for assessing their dependence. The difference between the sum of the marginal entropies and the joint entropy is given by

$$
J(X,Y) = H(X) + H(Y) - H(X,Y),
$$

which is a non-negative measure of association between $X$ and $Y$. It is equal to $0$ if and only if $X$ and $Y$ are stochastically independent, $X \bot Y$, that is, when $p(x,y) = p(x,+)p(+,y)$.

For three variables, the trivariate entropy is bounded by

$$
H(X,Y) \leq H(X,Y,Z) \leq H(X,Z) + H(Y,Z) - H(Z).
$$

The deviation from the upper bound is given by

$$
EJ(X,Y \mid Z) = H(X,Z) + H(Y,Z) - H(Z) - H(X,Y,Z),
$$

which is again non-negative and equals $0$ if and only if $X \bot Y \mid Z$. This quantity can therefore be used to assess conditional independence.

Joint entropy values $J(X,Y)$ for all pairs of variables can be used to explore the dependence structure through association graphs. In such graphs, nodes represent variables and edges connect pairs of variables whose joint entropy exceeds a chosen threshold. By gradually lowering the threshold, additional edges appear and the graph becomes more connected. 

Connected components that form cliques indicate groups of mutually dependent variables, while separate components correspond to independent subsets. Conditional independence can be investigated by examining how the graph structure changes when conditioning variables are removed. A threshold that yields a graph with several small (conditionally) independent components may serve as a useful representation of the underlying multivariate structure.

The function `joint_entropy()` computes joint entropy values for all pairs of variables in a data frame. The output is returned as a list containing: (i) an upper triangular matrix of joint entropy values, with univariate entropies on the diagonal, and (ii) a data frame summarizing the frequency distribution of unique joint entropy values. An optional argument controls the precision (number of decimal places) used in constructing this frequency distribution; the default is 3.

#### Vertex Variables {.unnumbered}

We apply the function to the data frame containing the observed and transformed vertex variables:

```{r}
#| label: vert-joint
j_vertex <- joint_entropy(vertex_var, 3)
```
The output is returned as a list containing two objects: `j_vertex$mat`, an upper triangular matrix of joint entropy values (with univariate entropies on the diagonal), and `j_vertex$freq`, a data frame summarizing the frequency distribution of unique joint entropy values.

The matrix `j_vertex$mat` shown in @tbl-joint-vert summarizes the pairwise joint entropy values. Entries equal to $0$ indicate independence between variables, while larger values indicate stronger dependence. For example, practice is independent of both status and gender. 

::: {#tbl-joint-vert}
```{r}
#| echo: false
knitr::kable(
  round(j_vertex$mat, 3),
)
```
Upper triangular matrix of joint entropy values for vertex variables, with univariate entropies on the diagonal.
:::

The strongest associations can be identified by combining the matrix with the frequency table `j_vertex$freq`, which summarizes the distribution of joint entropy values. This is shown in @tbl-jointfrq-vert where we see that the largest joint entropy value is $0.58$, occurring between the variables `years` and `status`.

::: {#tbl-jointfrq-vert}
```{r}
#| echo: false
knitr::kable(
  head(j_vertex$freq, 10)
)
```
Frequency distribution of unique joint entropy values for vertex variables. Only the first 10 rows are shown.
:::
  
These relationships can be conveniently visualized using the function `assoc_graph()`, which returns a `ggraph` object where nodes represent variables and edges represent the strength of association (with thicker edges indicating stronger dependence). To focus on the most relevant structure, we set a threshold of $0.25$, so that only the strongest associations are displayed:

```{r}
#| label: vert-assoc
#| fig-height: 8
#| fig-width: 10
assoc_graph(vertex_var, 0.25)
```

#### Dyad Variables {.unnumbered}

Joint entropies for the dyad variables are computed in the same way as for the vertex variables. Rather than presenting the full output, we focus on visualizing the dependence structure using an association graph. We set a threshold of 0.10 to display the strongest associations:

```{r}
#| label: dyad-assoc
#| fig-height: 8
#| fig-width: 10
assoc_graph(dyad_var, 0.10)
```

From the graph, several patterns can be observed. There is a strong dependence among the variables (`status`, `years`, `age`), as well as notable associations among (`lawschool`, `years`, `age`) and (`status`, `years`, `gender`). 

In addition, the graph suggests that the relations `cowork` and `friend` are conditionally independent given `advice`. That is, any observed dependence between `cowork` and `friend` can be explained through their association with `advice`.

#### Triad Variables {.unnumbered}

Similarly, joint entropies for the triad variables are computed for the reduced data frame (with redundant variables removed). The dependence structure is again visualized using an association graph. We set a threshold of 0.25 to focus on the strongest associations:

```{r}
#| label: triad-assoc
#| fig-height: 8
#| fig-width: 10
assoc_graph(triad_var, 0.25)
```
The triad variables exhibit similar dependence patterns as the dyad variables, but with generally stronger associations. In particular, joint entropy values are higher for all pairs of `cowork`, `friend`, and `advice`, indicating stronger interactions among these variables. This further supports the conditional independence observed earlier, suggesting that the relationship between `cowork` and `friend` is increasingly explained by `advice` at the triadic level.

### Trivariate Entropies and Prediction Power
Trivariate entropy extends the concepts of univariate and bivariate entropy to three variables. For variables $X$, $Y$, and $Z$, it is defined as
$$
H(X,Y,Z) = \sum_x \sum_y \sum_z p(x,y,z) \log_2 \frac{1}{p(x,y,z)},
$$
and satisfies the bounds
$$
H(X,Y) \leq H(X,Y,Z) \leq H(X,Z) + H(Y,Z) - H(Z).
$$

These bounds provide a basis for assessing dependence structures among triples of variables. In particular, the deviation from the lower bound is given by the expected conditional entropy
$$
EH(Z \mid X,Y) = H(X,Y,Z) - H(X,Y),
$$
which measures the remaining uncertainty in $Z$ given $(X,Y)$. This quantity is non-negative and equals $0$ if and only if $(X,Y)$ uniquely determine $Z$, indicating a functional relationship.

This concept parallels the bivariate case, where
$$
EH(Y \mid X) = H(X,Y) - H(X)
$$
measures the remaining uncertainty in $Y$ given $X$, and equals $0$ when $X$ fully determines $Y$.

More generally, $EH(Z \mid X,Y)$ can be interpreted as a measure of prediction uncertainty when $(X,Y)$ are used to predict $Z$. It reflects, on a logarithmic scale, the average number of possible outcomes of $Z$ given $(X,Y)$. Smaller values indicate stronger predictive power: if $EH < 0.5$, $Z$ can be predicted unambiguously; if $0.5 \leq EH < 1.5$, there are approximately two possible outcomes on average; and so on. In general, prediction power decreases as $EH$ increases.

The function `prediction_power()` computes this quantity for all pairs of variables in a given data frame used to predict a specified target variable $Z$. The function takes as input the data frame and the variable to be predicted, and returns an upper triangular matrix of expected conditional entropies $EH(Z \mid X,Y)$ for all pairs $(X,Y)$. The diagonal entries correspond to $EH(Z \mid X)$, that is, prediction based on a single variable. Entries corresponding to the target variable itself are set to `NA`.

For easier interpretation, prediction results can be visualized using prediction plots, which display a color-coded matrix with rows representing $X$ and columns representing $Y$. Darker colors indicate lower prediction uncertainty and thus higher predictive power for $Z$.
In the following, we compute and visualize prediction power for vertex, dyad, and triad variables.

#### Vertex Variables {.unnumbered}
Trivariate entropies for all combinations of vertex variables can be computed using the function `entropy_trivar()`:

```{r}
#| label: trivent_vert
h3_vertex <- entropy_trivar(vertex_var)
```

We next assess prediction power for the different degree-based variables derived from the three relation types: co-work, advice, and friendship. Specifically, we evaluate how well pairs of variables predict each of these target variables.

The corresponding prediction power matrices are obtained as follows:
```{r}
#| label: pred-power-vertex
pred_vertex_cwk    <- prediction_power("deg_cwk", vertex_var)
pred_vertex_advin  <- prediction_power("indeg_adv", vertex_var)
pred_vertex_advout <- prediction_power("outdeg_adv", vertex_var)
pred_vertex_advmut <- prediction_power("deg_mutadv", vertex_var)
pred_vertex_frnin  <- prediction_power("indeg_frn", vertex_var)
pred_vertex_frnout <- prediction_power("outdeg_frn", vertex_var)
pred_vertex_frnmut <- prediction_power("deg_mutfrn", vertex_var)
pred_vertex_frnlik <- prediction_power("deg_likfrn", vertex_var)
```
As an illustration, consider predicting the variable `indeg_adv`, which represents the in-degree in the advice network (i.e., how often a lawyer is asked for advice). The prediction power was computed an assigned to `pred_vertex_advin` and shown in @tbl-expred.


::: {#tbl-expred}
```{r}
#| echo: false
knitr::kable(
  round(pred_vertex_advin, 2),
)
```
Prediction power matrix for `indeg_adv`. Entries represent expected conditional entropies $EH(Z \mid X,Y)$, where lower values indicate higher predictive power. Diagonal entries correspond to single predictors, while off-diagonal entries represent prediction based on pairs of variables; `NA` values indicate the target variable itself.
:::



These prediction strengths can be visualized using a prediction plot in a heat map style. The plot displays a color-coded matrix with rows corresponding to $X$ and columns to $Y$, where darker colors indicate higher predictive power (i.e., lower prediction uncertainty for `indeg_adv`). This is shown in @fig-expred.


::: {#fig-expred}
```{r}
#| label: pred_heatmap
#| echo: false
#| fig-height: 7
#| fig-width: 8


# convert matrix to long format (base R)
mat <- as.matrix(pred_vertex_advin)

df <- as.data.frame(as.table(mat))
names(df) <- c("X", "Y", "EH")

# remove NA entries (target variable row/column)
df <- df[!is.na(df$EH), ]

ggplot(df, aes(x = Y, y = X, fill = EH)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(EH, 2)), size = 3) +
  scale_fill_gradient(
    low = "steelblue", 
    high = "white",
    name = "EH(Z|X,Y)"
  ) +
  labs(
    x = "Predictor Y",
    y = "Predictor X",
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

Prediction power heatmap for $Z$=`indeg_adv`. Each cell shows the expected conditional entropy $EH(Z \mid X,Y)$, where darker colors indicate stronger predictive power. The strongest predictors involve the variable `years`, particularly in combination with network-based variables such as `deg_mutadv` and `deg_cwk`. While `status` also contributes to prediction, it is not part of the strongest predictor pairs.
:::

The prediction power heatmap shows that the strongest overall predictor of `indeg_adv` is the network-based variable `deg_likfrn`, which yields the lowest expected conditional entropy among single predictors. This indicates that likely friendship ties are highly informative for explaining advice in-degree.

Among other single predictors, `years` also performs well, followed by `status`, while variables such as `office` and `practice` provide weaker predictive power.

Considering pairs of predictors, the strongest combinations typically involve `years` together with network-based variables such as `deg_mutadv` and `deg_cwk`, yielding the lowest prediction uncertainty overall.

These results suggest that both network position, particularly likely friendship ties, and experience (`years`) play a central role in explaining variation in advice in-degree.


```{r}
#| fig-height: 22
#| fig-width: 14
make_pred_plot <- function(mat, title) {
  df <- as.data.frame(as.table(as.matrix(mat)))
  names(df) <- c("X", "Y", "EH")
  
  df <- df[!is.na(df$EH), ]
  
  ggplot(df, aes(x = Y, y = X, fill = EH)) +
    geom_tile(color = "white") +
    geom_text(aes(label = round(EH, 2)), size = 2.5) +
    scale_fill_gradient(low = "steelblue", high = "white") +
    labs(title = title, x = NULL, y = NULL) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
      axis.text.y = element_text(size = 7),
      plot.title = element_text(size = 10)
    )
}
  
p1 <- make_pred_plot(pred_vertex_cwk,    "deg_cwk")
p2 <- make_pred_plot(pred_vertex_advin,  "indeg_adv")
p3 <- make_pred_plot(pred_vertex_advout, "outdeg_adv")
p4 <- make_pred_plot(pred_vertex_advmut, "deg_mutadv")
p5 <- make_pred_plot(pred_vertex_frnin,  "indeg_frn")
p6 <- make_pred_plot(pred_vertex_frnout, "outdeg_frn")
p7 <- make_pred_plot(pred_vertex_frnmut, "deg_mutfrn")
p8 <- make_pred_plot(pred_vertex_frnlik, "deg_likfrn")


(p1 | p2) /
(p3 | p4) /
(p5 | p6) /
(p7 | p8) 
```

