# Signed Networks {#sec-signed-networks}

Most of social network analysis deals with relations that are inherently
positive: friendship, collaboration, advice seeking, trade. The tools we use to
study networks, from centrality indices to community detection, were largely
developed with these positive ties in mind. But social life is not only about
who we like. People form rivalries, nations go to war, and online communities
like *and* dislike content. When we want to study networks that contain both
positive and negative ties, we enter the topic of signed networks.

Signed networks require their own analytical toolkit. Standard centrality
measures do not account for the difference between a node connected to five
friends versus one connected to five enemies. Community detection algorithms
that look for dense subgroups miss the point when the structure is defined by
who is *against* whom, not just who is *with* whom. In this chapter, we
introduce the key concepts and methods for analysing signed networks: structural
balance, blockmodeling, and centrality.

## Packages Needed for this Chapter

```{r}
#| label: libraries
#| message: false

library(igraph)
library(signnet)
library(networkdata)
```

```{r}
#| label: libraries-silent
#| include: false

library(ggraph)
library(patchwork)
```

## Basic Tools for Signed Networks

The `signnet` package provides dedicated tools for analysing signed networks in
R. The package builds on `igraph` and assumes that a signed network is an igraph
object with an edge attribute `sign`, containing values 1 (positive) or -1
(negative). All functions in the package check for this attribute and will throw
an error if it is missing or contains other values.

We will use the `tribes` dataset to illustrate some basic functionality of the
package. This dataset is a signed social network of sixteen tribes from the
Gahuku-Gama alliance structure of the Eastern Central Highlands of New Guinea.
Tribes are connected by friendship ties ("rova") and enmity ties ("hina").

```{r}
#| label: load-tribes

data("tribes")
```

The function `ggsigned()` provides a convenient way to visualise signed
networks. @fig-tribes shows the `tribes` network visualised with `ggsigned()`,
where blue edges represent positive ties and red edges represent negative ties.\

```{r}
#| label: fig-tribes
#| fig-cap: "The `tribes` network visualised with `ggsigned()`."

ggsigned(tribes)
```

The color of the edges can be customised with the `edge_cols` parameter.

The parameter `weights` can be used to artificially "pull" nodes connected with
negative ties apart, which can help to visually separate potential groups of
positive ties as shown in @fig-tribes-weighted.

```{r}
#| label: fig-tribes-weighted
#| fig-cap: "The `tribes` network visualised with `ggsigned()` using weights to separate negative ties."

ggsigned(tribes, weights = TRUE)
```

A signed network is represented by a signed adjacency matrix, where entries are
1, -1, or 0. It can be extracted with the function `as_adj_signed()`.

```{r}
#| label: signed-adj

as_adj_signed(tribes)
```

## Structural Balance

Structural balance theory originates in the social psychology of Fritz Heider
[-@heider1946attitudes] and was formalised for graphs by
@cartwright1956structural. The core idea is intuitive: consider three people. If
Alice and Bob are friends, and Bob and Carol are friends, we expect Alice and
Carol to also be friends: "the friend of my friend is my friend." Similarly, if
Alice and Bob are enemies and Bob and Carol are enemies, we expect Alice and
Carol to be friends: "the enemy of my enemy is my friend." These configurations
are **balanced**. A triangle where all three pairs are enemies, or where two
pairs are friends but one pair are enemies, feels unstable and is considered
**unbalanced**.

More formally, a triangle is balanced if it has an even number of negative ties
(zero or two), and unbalanced if it has an odd number (one or three). All four
possible configurations are shown in @fig-balance-triples.

```{r}
#| label: fig-balance-triples
#| fig-cap: "The four possible signed triangles and their balancedness."
#| echo: false

make_signed_triangle <- function(signs) {
  g <- make_graph(c(1, 2, 2, 3, 1, 3), directed = FALSE)
  E(g)$sign <- signs
  g
}

plot_triangle <- function(g, title) {
  E(g)$col <- ifelse(E(g)$sign == 1, "#104E8B", "firebrick3")
  edge_labels <- ifelse(E(g)$sign == 1, "+", "\u2013")
  ggraph(g, layout = "manual", x = c(-1, 1, 0), y = c(0, 0, sqrt(3))) +
    geom_edge_link(
      aes(label = edge_labels, edge_colour = I(col)),
      edge_width = 1.2,
      label_size = 6,
      angle_calc = "across",
      label_push = unit(-3, "mm")
    ) +
    geom_node_point(size = 5, colour = "grey30") +
    theme_graph() +
    theme(plot.title = element_text(hjust = 0.5, size = 11))
}

t1 <- make_signed_triangle(c(1, 1, 1))
t2 <- make_signed_triangle(c(-1, -1, 1))
t3 <- make_signed_triangle(c(1, 1, -1))
t4 <- make_signed_triangle(c(-1, -1, -1))

p1 <- plot_triangle(t1, "+++") + 
  theme(axis.title.x = element_text(face = "bold",size = 12)) +
  scale_x_continuous(name = "balanced")
p2 <- plot_triangle(t2, "+\u2013\u2013") + 
  theme(axis.title.x = element_text(face = "bold",size = 12)) +
  scale_x_continuous(name = "balanced")
p3 <- plot_triangle(t3, "++\u2013") + 
  theme(axis.title.x = element_text(face = "bold",size = 12)) +
  scale_x_continuous(name = "unbalanced")
p4 <- plot_triangle(t4, "\u2013\u2013\u2013") + 
  theme(axis.title.x = element_text(face = "bold",size = 12)) +
  scale_x_continuous(name = "unbalanced")


p1 + p2 + p3 + p4 + plot_layout(ncol = 2, axis = "collect_x")

```

A network is **balanced** if it can be partitioned into two groups such that all
ties within groups are positive and all ties between groups are negative.

We can generate such a simple balanced network with `sample_islands_signed()`,
which creates groups with positive intra-group ties and negative inter-group
ties. It follows the logic of the function `sample_islands()` from `igraph`,
with the added twist of making inter group edges negative. Setting
`islands.n = 2` creates a perfectly balanced network with two groups.

```{r}
#| label: balanced-example

g_bal <- sample_islands_signed(
  islands.n = 2, 
  islands.size = 10, 
  islands.pin = 0.8, 
  n.inter = 5
)
```

The network is shown in @fig-balanced.

```{r}
#| label: fig-balanced
#| fig-cap: "A synthetic perfectly balanced network with two groups."

ggsigned(g_bal)
```

We can verify that this network is perfectly balanced by counting its triangles.
Balanced networks contain only "+++" and "+--" triangles, with no "++-" or "---"
triangles.

```{r}
#| label: count-triangles

count_signed_triangles(g_bal)
```

Increasing the `islands.n` parameter to more than two creates networks that are
not balanced in the strict two-group sense but are "clusterable" as defined by
@davis1967clustering -- they can be partitioned into more than two groups with
the same within-positive, between-negative pattern.

Real-world networks are rarely perfectly balanced. The question then becomes:
how balanced a given network is. The `signnet` package offers three methods for
measuring the degree of balancedness via `balance_score()`. All return a value
between zero (perfectly unbalanced) and one (perfectly balanced).

For our perfectly balanced synthetic network, all three methods obviously agree.

```{r}
#| label: balance-perfect

balance_score(g_bal, method = "triangles")
balance_score(g_bal, method = "walk")
balance_score(g_bal, method = "frustration")
```

The **triangles** method returns the fraction of balanced triangles. It is the
most intuitive but only considers local structure.

The **walk** method uses eigenvalues of the signed and unsigned adjacency
matrices to capture balance at all scales, not just triangles. It is based on
the work of @estrada2019rethinking.

The **frustration** method finds a partition of the network into two groups and
counts how many edges violate the balance pattern (negative ties within groups
and positive ties between groups). The fewer violations, the more balanced the
network [@aref2018measuring]. Since finding the optimal partition is
computationally hard, the function uses simulated annealing and returns an upper
bound. For exact results, `frustration_exact()` solves the problem via integer
programming for small networks. For the `tribes` network, which is not perfectly
balanced, the three methods give the following results.

```{r}
#| label: balance-tribes

balance_score(tribes, method = "triangles")
balance_score(tribes, method = "walk")
balance_score(tribes, method = "frustration")
```

This disagreement is common for empirical networks and reflects the fact that
each method captures different aspects of balance.

## Blockmodeling

Structural balance theory dictates that a balanced network can be partitioned
into groups with positive intra-group ties and negative inter-group ties.
Blockmodeling is a computational approach to finding such partitions
[@doreian1996partitioning].

The function `signed_blockmodel()` partitions a network into `k` predefined
blocks, optimising for positive ties within blocks and negative ties between
blocks. The objective function to be optimized is
$P(C) = \alpha N + (1-\alpha)P$, where $N$ is the total number of negative ties
within blocks and $P$ is the total number of positive ties between blocks. The
parameter `alpha` controls the trade-off between these two values. The closer
`alpha` is to 1, the more the algorithm prioritises negative ties within blocks.
The closer to 0, the more it prioritises positive ties between blocks. Setting
`alpha = 0.5` gives equal weight to both types of errors.

For the `tribes` network, we try to find a partition into three blocks, which is
the known structure of the Gahuku-Gama alliance system.

```{r}
#| label: blockmodel-tribes

set.seed(1108)
clu <- signed_blockmodel(tribes, k = 3, alpha = 0.5, annealing = TRUE)
clu$membership
clu$criterion
```

The function returns a list with two entries: the block membership of each node
and the value of $P(C)$, where lower values indicate a better fit. Setting
`annealing = TRUE` uses simulated annealing in the optimisation, which generally
produces better results at the cost of longer running time. The result shows
that only `r clu$criterion` edges are either positive intergroup ties or
negative intragroup ties.

The obtained block structure can be visualized with `ggblock()` as shown in
@fig-blockmodel-tribes.

```{r}
#| label: fig-blockmodel-tribes
#| fig-cap: "Block structure of the `tribes` network."

ggblock(tribes, clu$membership, show_blocks = TRUE)
```

The figure shows that the two edges that violate the balance pattern are between
groups two and three, which have two positive ties between them. The rest of the
structure is perfectly balanced, with positive ties within groups and negative
ties between groups.

The traditional blockmodeling assumes a single pattern of positive blocks on the
diagonal and negative blocks off the diagonal. But many networks have more
complex structures [@doreian2009partitioning]. For instance, two groups might be
allies (positive ties between them) while both are enemies of a third group.

The function `signed_blockmodel_general()` allows us to specify arbitrary block
structures via the `blockmat` parameter. Each entry in the matrix indicates
whether we expect the corresponding block to be positive (1) or negative (-1).

In the following we construct a synthetic network with three groups, with a
structure which cannot be captured by the traditional blockmodel but can be
specified in the general blockmodel.

```{r}
#| label: general-setup

## construct a network with a non-standard block structure
g1 <- g2 <- g3 <- make_full_graph(5)
V(g1)$name <- as.character(1:5)
V(g2)$name <- as.character(6:10)
V(g3)$name <- as.character(11:15)

g <- Reduce("%u%", list(g1, g2, g3))
E(g)$sign <- 1
E(g)$sign[1:10] <- -1
g <- add_edges(g, c(rbind(1:5, 6:10)), attr = list(sign = -1))
g <- add_edges(g, c(rbind(1:5, 11:15)), attr = list(sign = -1))
g <- add_edges(g, c(rbind(11:15, 6:10)), attr = list(sign = 1))
```

```{r}
#| label: fig-general-setup
#| fig-cap: "A synthetic network with a non-standard block structure."

ggsigned(g, weights = TRUE)
```

Here, groups two and three have positive ties between them, while both have
negative ties with group one. The block structure matrix specified below
reflects this.

```{r}
#| label: general-blockmat

set.seed(1108)
blockmat <- matrix(
  c(1, -1, -1,
   -1, 1, 1,
   -1, 1, -1),
    3, 3, byrow = TRUE)

blockmat

general <- signed_blockmodel_general(g, blockmat, alpha = 0.5)
general$criterion
```

The obtained block structure is shown in @fig-general-blockmodel.

```{r}
#| label: fig-general-blockmodel
#| fig-cap: "Block structure of the synthetic network with a non-standard block structure."

ggblock(g, general$membership, show_blocks = TRUE)
```

Comparing the general model with the traditional model on this network shows the
benefit of specifying the correct block structure.

```{r}
#| label: compare-blockmodels

traditional <- signed_blockmodel(g, k = 3, alpha = 0.5, annealing = TRUE)
c(general = general$criterion, traditional = traditional$criterion)
```

A lower criterion value indicates a better fit. The general model performs
better here because it matches the actual structure of the network.

## Centrality

In unsigned networks, having many connections is straightforwardly advantageous.
But in signed networks, the picture is more nuanced. A tribe with six alliances
and no enemies is in a very different position from one with three alliances and
three enemies, even though both have the same number of ties. Centrality indices
for signed networks must account for this distinction.

The function `degree_signed()` offers four variants, controlled by the `type`
parameter:

- `type = "pos"`: count only positive neighbors
- `type = "neg"`: count only negative neighbors
- `type = "ratio"`: positive neighbors / (positive + negative neighbors)
- `type = "net"`: positive neighbors minus negative neighbors

For directed networks, the `mode` parameter distinguishes "in" and "out"
versions.

```{r}
#| label: degree-tribes

data.frame(
  tribe = V(tribes)$name,
  pos = degree_signed(tribes, type = "pos"),
  neg = degree_signed(tribes, type = "neg"),
  ratio = round(degree_signed(tribes, type = "ratio"), 2),
  net = degree_signed(tribes, type = "net")
)
```

Beyond degree, the `signnet` package implements two more centrality indices. The
**PN index** by @everett2014networks is conceptually similar to Katz status for
unsigned networks. It takes into account not just direct ties but also the signs
of indirect paths through the network.

**Eigenvector centrality** (`eigen_centrality_signed()`) extends the idea that a
node is central if it is connected to other central nodes, adapted for signed
networks [@bonacich2004calculating].

```{r}
#| label: centrality-comparison

cent_df <- data.frame(
  tribe = V(tribes)$name,
  degree_net = degree_signed(tribes, type = "net"),
  eigen = round(eigen_centrality_signed(tribes), 3),
  pn = round(pn_index(tribes), 3)
)
cent_df
```

The correlation between eigenvector centrality and the PN index tells us how
much these two measures agree.

```{r}
#| label: centrality-cor

cor(eigen_centrality_signed(tribes), pn_index(tribes), method = "kendall")
```

Note that the adjacency matrix of a signed network may not have a dominant
eigenvalue. When this occurs, eigenvector centrality is not well-defined and
`eigen_centrality_signed()` will return an error. In such cases, the PN index
provides a more robust alternative.

## Signed Two-Mode Networks

The `signnet` package also provides tools for working with signed two-mode
networks. Projecting a signed two-mode network onto one of its modes is less
straightforward than in the unsigned case, because paths through negative ties
can change the sign of the resulting projection. The package implements a
duplication approach that resolves this issue and also accounts for ambivalent
ties, where both positive and negative paths exist between two nodes. For
details, see the `signnet` package vignette on signed two-mode networks.

## Use Case: International Relations

The `cowList` dataset (included in the `signnet` package; see
@doreian2015structural) contains 51 signed networks of inter-state relations in
overlapping four-year windows from 1946 to 1999, derived from the [Correlates of
War](https://correlatesofwar.org/) project. Two countries are connected by a
positive tie if they formed an alliance or signed a peace treaty, and by a
negative tie if they were at war or involved in other conflicts.

```{r}
#| label: load-cow

data("cowList")
names(cowList)[c(1, 20, 51)]
```

Each network covers a four-year window (e.g., "65-68" covers 1965--1968). Let us
examine two snapshots: one from the height of the Cold War and one from the
post-Cold War period.

```{r}
#| echo: false

cowList <- lapply(cowList, igraph::upgrade_graph)
```
```{r}
#| label: cow-years

cow_cold <- cowList[["65-68"]]
cow_post <- cowList[["93-96"]]
```

```{r}
#| label: cow-sizes

c(nodes_cold = vcount(cow_cold), edges_cold = ecount(cow_cold),
  nodes_post = vcount(cow_post), edges_post = ecount(cow_post))
```

We can assess how balanced the international system was in each period:

```{r}
#| label: cow-balance

data.frame(
  method = c("triangles", "walk", "frustration"),
  cold_war = c(
    balance_score(cow_cold, method = "triangles"),
    balance_score(cow_cold, method = "walk"),
    balance_score(cow_cold, method = "frustration")
  ),
  post_cold_war = c(
    balance_score(cow_post, method = "triangles"),
    balance_score(cow_post, method = "walk"),
    balance_score(cow_post, method = "frustration")
  )
)
```

Applying blockmodeling reveals the alliance structure. During the Cold War, we
would expect to find blocs corresponding to the Western and Eastern alliances:

```{r}
#| label: cow-blockmodel

set.seed(1108)
clu_cold <- signed_blockmodel(cow_cold, k = 2, alpha = 0.5, annealing = TRUE)
```

```{r}
#| label: plot-cow-blockmodel

ggblock(cow_cold, clu_cold$membership, show_blocks = TRUE)
```

The block membership can be inspected alongside the country names to see which
states cluster together:

```{r}
#| label: cow-membership

split(V(cow_cold)$name, clu_cold$membership)
```
