# Enhancing Visualizations

Everything that was covered so far should be enough to produce standard high-quality network visualizations. However, `ggraph` offers a lot more advanced functions and parameter settings to further enhance network plots. 


## Packages Needed for this Chapter
```{r}
#| label: libraries1
#| message: false
library(igraph)
library(ggraph)
library(graphlayouts)
library(ggforce)
library(networkdata)
```

As before, we use the Game of Thrones dataset, which is part of the `networkdata` package.

```{r}
data("got")

gotS1 <- got[[1]]

got_palette <- c(
  "#1A5878",
  "#C44237",
  "#AD8941",
  "#E99093",
  "#50594B",
  "#8968CD",
  "#9ACD32"
)

## compute a clustering for node colors
V(gotS1)$clu <- as.character(membership(cluster_louvain(gotS1)))

## compute degree as node size
V(gotS1)$size <- degree(gotS1)
```

## Combining `ggraph` and `ggforce`

The `ggforce` package works and `ggraph` work pretty nicely together. For example, you can use
the `geom_mark_*()` functions to highlight clusters.

```{r}
#| label: network-grps
set.seed(665)

## create network with a group structure
g <- sample_islands(9, 40, 0.4, 15)
g <- igraph::simplify(g)
V(g)$grp <- as.character(rep(1:9, each = 40))

```

```{r}
#| label:  fig-network-grps-sol
#| fig-cap: "Network with group structure highlighted using `geom_mark_hull()`."
ggraph(g, layout = "backbone", keep = 0.4) +
  geom_edge_link0(edge_color = "grey66", edge_width = 0.2) +
  geom_node_point(aes(fill = grp), shape = 21, size = 3) +
  geom_mark_hull(
    aes(x, y, group = grp, fill = grp),
    concavity = 4,
    expand = unit(2, "mm"),
    alpha = 0.25
  ) +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  scale_edge_color_manual(values = c(rgb(0, 0, 0, 0.3), rgb(0, 0, 0, 1))) +
  theme_graph() +
  theme(legend.position = "none")
```

@fig-network-grps-sol shows the result. It is also possible to add labels to the clusters.

```{r}
#| label: fig-network-grps-label-sol
#| fig-cap: "Network with labelled group hulls."
ggraph(g, layout = "backbone", keep = 0.4) +
  geom_edge_link0(edge_color = "grey66", edge_width = 0.2) +
  geom_node_point(aes(fill = grp), shape = 21, size = 3) +
  geom_mark_hull(
    aes(x, y, group = grp, fill = grp, label = grp),
    concavity = 4,
    expand = unit(2, "mm"),
    alpha = 0.25
  ) +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  scale_edge_color_manual(values = c(rgb(0, 0, 0, 0.3), rgb(0, 0, 0, 1))) +
  theme_graph() +
  theme(legend.position = "none")
```

The labelled version is shown in @fig-network-grps-label-sol.

## Small tricks for common problems 

> "How can I achieve that my directed edges stop at the node border, independent from the node size?" 

This one has given me headaches for the longest time. No matter what I tried, I always ended up with
something like @fig-arrow-size.

```{r}
#| label: fig-arrow-size
#| fig-cap: "Directed network where arrows overlap with nodes of varying size."
## create a random network
set.seed(1071)
g <- sample_pa(30, 1)
V(g)$degree <- degree(g, mode = "in")

ggraph(g, "stress") +
  geom_edge_link(
    aes(end_cap = circle(node2.degree + 2, "pt")),
    edge_colour = "black",
    arrow = arrow(
      angle = 10,
      length = unit(0.15, "inches"),
      ends = "last",
      type = "closed"
    )
  ) +
  geom_node_point(aes(size = degree), col = "grey66", show.legend = FALSE) +
  scale_size(range = c(3, 11)) +
  theme_graph()
```

The overlap can be avoided by using the `I()` function from base R, which
treats the entries of a vector "as is". So we know that if a node has degree 5, it will be mapped to
a circle with radius (or diameter?) "5pt". Since this means, that you have no control over the scaling,
you need to do that beforehand.

```{r}
#| label: fig-arrows-size-sol
#| fig-cap: "Directed network with corrected arrow endpoints using `I()` for node sizes."
## this function is borrowed from the ambient package
normalise <- function(x, from = range(x), to = c(0, 1)) {
  x <- (x - from[1]) / (from[2] - from[1])
  if (!identical(to, c(0, 1))) {
    x <- x * (to[2] - to[1]) + to[1]
  }
  x
}

## map to the range you want
V(g)$degree <- normalise(V(g)$degree, to = c(3, 11))

ggraph(g, "stress") +
  geom_edge_link(
    aes(end_cap = circle(node2.degree + 2, "pt")),
    edge_colour = "grey25",
    arrow = arrow(
      angle = 10,
      length = unit(0.15, "inches"),
      ends = "last",
      type = "closed"
    )
  ) +
  geom_node_point(aes(size = I(degree)), col = "grey66") +
  theme_graph()
```

The corrected version is shown in @fig-arrows-size-sol. I would not be surprised though if there is an even easier fix for this problem.

> "How can I lower the opacity of nodes without making edges visible underneath?"

One of the rules I try to follow is that edges should not be visible on top of nodes.
Usually that is easy to achieve by drawing the edges before the nodes. But if you
want to lower the opacity of nodes, they do become visible again, as shown in @fig-alpha-nodes.

```{r}
#| label: fig-alpha-nodes
#| fig-cap: "Network with transparent nodes where edges are visible underneath."
g <- sample_gnp(20, 0.5)
V(g)$degree <- degree(g)

ggraph(g, "stress") +
  geom_edge_link(edge_colour = "grey66") +
  geom_node_point(
    size = 8,
    aes(alpha = degree),
    col = "red",
    show.legend = FALSE
  ) +
  theme_graph()
```

The solution is rather simple. Just add a node layer with the same aesthetics below with
`alpha=1` (default) and `color="white"` (or the background color of the plot).
```{r}
#| label: fig-alpha-nodes-sol
#| fig-cap: "Network with opaque white background layer to hide edges beneath transparent nodes."
ggraph(g, "stress") +
  geom_edge_link(edge_colour = "grey66") +
  geom_node_point(size = 8, col = "white") +
  geom_node_point(
    aes(alpha = degree),
    size = 8,
    col = "red",
    show.legend = FALSE
  ) +
  theme_graph()
```

The result is shown in @fig-alpha-nodes-sol. Of course you could also use `start_cap` and `end_cap` here, but you may have to fiddle again as in the last example. 

> "How can I enhance readability of node labels in hairball graphs?"

Sometimes it is really hard to make labels readable when the network is very cluttered (see @fig-unreadable-labels).

```{r}
#| label: fig-unreadable-labels
#| fig-cap: "Dense network with hard-to-read node labels."
#| fig-width: 8
#| fig-height: 8
g <- sample_gnp(50, 0.7)
V(g)$name <- sapply(1:50, function(x) paste0(sample(LETTERS, 4), collapse = ""))
E(g)$weight <- runif(ecount(g))

ggraph(g) +
  geom_edge_link0(
    aes(edge_color = weight, edge_linewidth = weight),
    show.legend = FALSE
  ) +
  geom_node_point(size = 8, color = "#44a6c6") +
  geom_node_text(aes(label = name), fontface = "bold") +
  scale_edge_color_continuous(low = "grey66", high = "black") +
  scale_edge_width(range = c(0.1, 0.5)) +
  theme_graph() +
  coord_fixed()
```

Here you can make use of the fact that the layout of the nodes are stored in a "hidden" data frame when a `ggraph` object is constructed. That means you can use other geoms from other packages. In this case, the `shadowtext` package as shown in @fig-shadowtext-labels.

```{r}
#| label: fig-shadowtext-labels
#| fig-cap: "Dense network with improved label readability using `shadowtext`."
#| fig-width: 8
#| fig-height: 8

ggraph(g, "stress") +
  geom_edge_link0(
    aes(edge_color = weight, edge_linewidth = weight),
    show.legend = FALSE
  ) +
  geom_node_point(size = 8, color = "#44a6c6") +
  shadowtext::geom_shadowtext(
    aes(x, y, label = name),
    color = "black",
    size = 4,
    bg.colour = "white"
  ) +
  scale_edge_color_continuous(low = "grey66", high = "black") +
  scale_edge_width(range = c(0.1, 0.5)) +
  theme_graph() +
  coord_fixed()
```
