# A Continuous-Time Markov Chain (Cat Behavior Model) {#sec-apx-cat .appendix}

```{r}
#| label: libraries-silent
#| include: false

library(igraph)
library(ggraph)
library(tidygraph)
library(ggplot2)
```

To bring the concept of a continuous-time Markov chain (CTMC) to life (or to all
nine lives of our cat), consider a model of a house cat's daily activities. At
any given moment, the cat is in one of the following behavioral states:

- $0$: Sleeping
- $1$: Eating
- $2$: Playing
- $3$: Plotting chaos (e.g., knocking things off shelves)

We define $X(t)$ as the cat's current activity at time $t$. The process
$\{X(t), t \in [0, \infty)\}$ satisfies:

- A finite state space $S = \{0, 1, 2, 3\}$\
- Continuous transitions over time\
- The Markov property: next state depends only on the current state\

The cat transitions between states at random times. Each stay in a state lasts
for a random holding time, and transitions to the next state occur after a random holding time, and the next
state is chosen probabilistically according to a transition (jump) matrix.

The holding time $T_i$ in state $i$ is modeled using an exponential
distribution:
$$
f_{T_i}(t) = \lambda_i e^{-\lambda_i t}, \quad t > 0
$$

- $\lambda_i$ is the rate of leaving state $i$.
- $\mathbb{E}[T_i] = \frac{1}{\lambda_i}$ is the expected duration in state $i$.

The exponential distribution's **memoryless property** means that the
probability of remaining in a state is independent of how long the cat has
already been in it:
$$
P(T_i > s + t \mid T_i > t) = P(T_i > s)
$$
So, even after two hours of napping, the chance that the cat naps another 30
minutes is the same as if it had just started.

Once the holding time ends, the cat jumps to a new state. The **transition
(jump) matrix** $P = (p_{ij})$ governs this:
$$
p_{ij} = P(X(t^+) = j \mid X(t) = i)
$$

For each state $i$, the row of probabilities $p_{ij}$ must sum to 1:
$$
\sum_{j \in S} p_{ij} = 1
$$

@fig-catex illustrates a single realization of such a process. This visual shows
how a process starting in state 0 might stay there for some time, then jump to
state 2, then state 3, and so on, with irregular intervals between jumps.

::: {#fig-catex}
```{r}
#| label: catex-plot
#| fig-height: 4
#| fig-width: 7
#| message: false
#| echo: false

jump_times <- c(0, 2, 5, 8, 14)
states <- c(0, 2, 3, 0, 1)  # Sleep → Play → Chaos → Sleep → Eat

segment_data <- data.frame(
  x = jump_times[-length(jump_times)],
  xend = jump_times[-1],
  y = states[-length(states)]
)

jump_points <- data.frame(
  x = jump_times[-1],
  y = states[-1]
)

ggplot() +
  geom_segment(data = segment_data, aes(x = x, xend = xend, y = y, yend = y),
               linewidth = 1.2, color = "black") +
  geom_point(data = jump_points, aes(x = x, y = y), shape = 21, fill = "black", size = 3) +
  annotate("segment", x = 0, xend = 2, y = -0.3, yend = -0.3,
           arrow = arrow(type = "closed", length = unit(0.2, "cm")),
           color = "#1c9099", linewidth = 1.5) +
  annotate("text", x = 1, y = -0.5, label = "Holding time", color = "#1c9099", size = 4) +
  annotate("segment", x = 8, xend = 8, y = 3, yend = 0.09,
           arrow = arrow(type = "closed", length = unit(0.2, "cm")),
           color = "darkorange", linewidth = 1.5) +
  annotate("text", x = 8.2, y = 1.5, label = "Jump", color = "darkorange", size = 4, angle = 90) +
  labs(x = "Time", y = "State") +
  scale_y_continuous(breaks = 0:3,
                     labels = c("Sleep", "Eat", "Play", "Chaos")) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        axis.title.y = element_text(angle = 0, vjust = 1.2))
```
A realization of a continuous-time Markov chain (CTMC) showing the cat
transitioning through behavioral states (Sleep → Play → Chaos → Sleep → Eat).
Horizontal segments show how long each activity lasts (holding time), while
vertical arrows show transitions (jumps) between states.

:::

Below is a hypothetical transition matrix $P$ for the cat's behavioral states.
Each row corresponds to the current state, and each column to the next state:

  |           | Sleep | Eat  | Play | Chaos |
  | --------- | :---: | :--: | :--: | :---: |
  | **Sleep** | 0.00  | 0.70 | 0.83 | 0.56  |
  | **Eat**   | 0.71  | 0.00 | 0.81 | 0.20  |
  | **Play**  | 0.05  | 0.82 | 0.00 | 0.45  |
  | **Chaos** | 0.91  | 0.46 | 0.61 | 0.00  |

> *Note: Diagonal entries (e.g., Sleep → Sleep) are set to zero for
> interpretability. They can be included to model the probability of no state
> change.*

Next, we can combine the states and transitions into a directed graph showing
which states can be reached from one another, and with what likelihood. This is
shown in @fig-catex2.

::: {#fig-catex2}
```{r}
#| label: catex2-plot
#| fig-height: 4
#| fig-width: 4
#| message: false
#| echo: false

# Define all 4 states
state_names <- c("Sleep", "Eat", "Play", "Chaos")

# Create all combinations of from → to (including self-loops)
transitions <- expand.grid(from = state_names, to = state_names, stringsAsFactors = FALSE)

# Assign random-ish transition probabilities (you can edit or normalize)
set.seed(1108)
transitions$prob <- round(runif(nrow(transitions), min = 0.05, max = 0.95), 2)
# Create graph
g <- tbl_graph(nodes = data.frame(name = state_names), edges = transitions, directed = TRUE)

# Plot using geom_edge_arc to support curvature
ggraph(g, layout = "circle") +
  geom_edge_arc(aes(label = prob),
                arrow = arrow(length = unit(3, "mm"), type = "closed"),
                end_cap = circle(8, "mm"),
                start_cap = circle(8, "mm"),
                label_color = "black",
                label_size = 3,
                alpha = 0.8,
                strength = 0.15,
                angle_calc = "along",
                label_dodge = unit(2.5, "mm"),
                show.legend = FALSE) +
  geom_node_circle(aes(r = 0.1), fill = "lightgrey", color = "black") +
  geom_node_text(aes(label = name), size = 3, fontface = "bold") +
  theme_void() 


```
Directed graph representation of the CTMC jump chain. Each node is a state
(e.g., Sleep, Eat), and arrows represent possible transitions with associated
probabilities.
:::

@fig-catex2 is a visual representation of this matrix. Each arrow in the graph
corresponds to a non-zero entry $p_{ij}$ in the matrix. The curved edges
indicate transitions between pairs of states, and the labels on the arrows match
the values in the matrix. Together, the matrix and the graph describe a **jump
chain** over the set of behavioral states. These transitions are stochastic
(i.e., random), and their dynamics unfold in continuous time, which is what
differentiates CTMCs from discrete-time Markov models.

Together, the holding time distribution and the transition (jump) matrix fully
characterize the dynamics of the CTMC.

In summary, a Continuous-Time Markov Chain (CTMC):

- Determines how long the system remains in a state using the holding time,
  typically modeled as exponentially distributed.
- Uses a transition (jump) matrix to govern which state is entered next.
- Is memoryless and evolves in continuous time, meaning the future depends only
  on the present state and not the past.

These principles underpin Stochastic Actor-Oriented Models (SAOMs), where actors
make sequential and probabilistic changes to their network ties or attributes,
driven solely by the current network state.
